# More `ggplot2` 

```{r, echo=FALSE}
# Unattach any packages that happen to already be loaded. In general this is unecessary
# but is important for the creation of the book to not have package namespaces
# fighting unexpectedly.
pkgs = names(sessionInfo()$otherPkgs)
if( length(pkgs > 0)){
  pkgs = paste('package:', pkgs, sep = "")
  for( i in 1:length(pkgs)){
    detach(pkgs[i], character.only = TRUE, force=TRUE)
  }
}
```

## Faceting

The goal with faceting is to make many panels of graphics where each panel represents the same relationship between variables, but something changes between each panel. For example using the `iris` dataset we could look at the relationship between `Sepal.Length` and `Petal.Length` either with all the data in one graph, or one panel per species.

```{r, fig.height=3}
library(ggplot2)
ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width)) +
  geom_point() +
  facet_grid( . ~ Species )
```

The line `facet_grid( formula )` tells `ggplot2` to make panels, and the formula tells how to orient the panels. Recall that a formula is always in the order `y ~ x` and because I want the species to change as we go across the page, but don't have anything I want to change vertically we use `. ~ Species` to represent that. If we had wanted three graphs stacked then we could use `Species ~ .`. 

For a second example, we look at a dataset that examines the amount a waiter was tipped by 244 parties. Covariates that were measured include the day of the week, size of the party, total amount of the bill, amount tipped, whether there were smokers in the group and the gender of the person paying the bill

```{r}
data(tips, package='reshape')
head(tips)
```

It is easy to look at the relationship between the size of the bill and the percent tipped.

```{r, fig.height=3}
ggplot(tips, aes(x = total_bill, y = tip / total_bill )) +
  geom_point()
```

Next we ask if there is a difference in tipping percent based on gender or day of the week by plotting this relationship for each combination of gender and day.

```{r, fig.height=3}
ggplot(tips, aes(x = total_bill, y = tip / total_bill )) +
  geom_point() +
  facet_grid( sex ~ day )
```

Sometimes we want multiple rows and columns of facets, but there is only one categorical variable with many levels. In that case we use facet_wrap which takes a one-sided formula.
```{r, fig.height=3}
ggplot(tips, aes(x = total_bill, y = tip / total_bill )) +
  geom_point() +
  facet_wrap( ~ day )
```


Finally we can allow the x and y scales to vary between the panels by setting “free”, “free_x”, or “free_y”. In the following code, the y-axis scale changes between the gender groups.
```{r, fig.height=3}
ggplot(tips, aes(x = total_bill, y = tip / total_bill )) +
  geom_point() +
  facet_grid( sex ~ day, scales="free_y" )
```


## Modifying Scales

Often it is useful to modify the scales that we have on the x or y axis. In particular we might want to display some modified version of a variable. 

### Log scales

For this example, we'll use the `ACS` data from the `Lock5Data` package that has information about `Income` (in thousands of dollars) and `Age`. Lets make a scatterplot of the data.
```{r, fig.height=3}
library(Lock5Data)
data(ACS)
ggplot(ACS, aes(x=Age, y=Income)) +
  geom_point()
```

This is an ugly graph because six observations dominate the graph and the bulk of the data (income < $100,000) is squished together. One solution is to plot income on the $\log_{10}$ scale. There are a couple ways to do this. The simplest way is to just do a transformation on the column of data.
```{r, fig.height=3}
ggplot(ACS, aes(x=Age, y=log10(Income))) +
  geom_point()
```


This works quite well to see the trend of peak earning happening in a persons 40s and 50s, but the scale is difficult for me to understand (what does $\log_{10}\left(X\right)=1$ mean here? Oh right, that is $10^{1}=X$ so that is the $10,000 line). It would be really nice if we could do the transformation but have the labels on the original scale. 
```{r, fig.height=3}
ggplot(ACS, aes(x=Age, y=Income)) +
  geom_point() +
  scale_y_log10()
```

Now the y-axis is in the original units (thousands of dollars) but obnoxiously we only have two labeled values. Lets define the major break points (the white lines that have numerical labels) to be at 1,10,100 thousand dollars in salary. Likewise we will tell `ggplot2` to set minor break points at 1 to 10 thousand dollars (with steps of 1 thousand dollars) and then 10 thousand to 100 thousand but with step sizes of 10 thousand, and finally minor breaks above 100 thousand being in steps of 100 thousand.
```{r}
ggplot(ACS, aes(x=Age, y=Income)) +
  geom_point() +
  scale_y_log10(breaks=c(1,10,100),
                minor=c(1:10,
                        seq( 10, 100,by=10 ),
                        seq(100,1000,by=100))) +
  ylab('Income (1000s of dollars)')
```


### Arbitrary transformations

The function `scale_y_log10()` is actually just a wrapper to the `scale_y_continuous()` function with a predefined transformation. If you want to rescale function using some other function (say the inverse, or square-root, or $\log_{2}$
 ) you can use the `scale_y_continuous()` function (for the x-axis there is a corresponding `scale_x_????`) family of functions. There is a whole list of transformations built into ggplot2 that work (transformations include "asn", "atanh", "boxcox", "exp", "identity", "log", "log10", "log1p", "log2", "logit", "probability", "probit", "reciprocal", "reverse" and "sqrt"). If you need a custom function, that can be done by defining a new transformation via the trans_new() function.

## Multi-plot

There are times that you must create a graphic that is composed of several sub-graphs and think of it as one object. Unfortunately the mechanism that `ggplot2` gives for this is cumbersome and it is usually easier to use a function called `multiplot`. The explaination I've heard about why this function wasn't included in ggplot2 is that you should think about faceting first and only resort to `multiplot` if you have to. The function `multiplot` is included in a couple of packages, e.g. `Rmisc`, but I always just google 'ggplot2 multiplot' to get to the webpage [http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/]

```{r}
# This example uses the ChickWeight dataset, which comes with ggplot2 
# First plot 
p1 <- ggplot(ChickWeight, aes(x=Time, y=weight, colour=Diet, group=Chick)) +
    geom_line() +
    ggtitle("Growth curve for individual chicks")

# Second plot
p2 <- ggplot(ChickWeight, aes(x=Time, y=weight, colour=Diet)) +
    geom_point(alpha=.3) +
    geom_smooth(alpha=.2, size=1) +
    ggtitle("Fitted growth curve per diet")

# Third plot 
p3 <- ggplot(subset(ChickWeight, Time==21), aes(x=weight, colour=Diet)) +
    geom_density() +
    ggtitle("Final weight, by diet")
```


Suppose that I want to layout these three plots in an arrangement like so:

\[\textrm{layout}=\left[\begin{array}{ccc}
1 & 2 & 2\\
1 & 2 & 2\\
1 & 3 & 3
\end{array}\right]\]

 where plot 1 is a tall, skinny plot on the left, plot 2 is more squarish, and plot 3 is short on the bottom right. This sort of table arrangement can be quite flexible if you have many rows and many columns, but generally we can get by with something with only a couple rows/columns.
```{r, fig.height=4, warning=FALSE, message=FALSE}
my.layout = cbind( c(1,1,1), c(2,2,3), c(2,2,3) )
Rmisc::multiplot( p1, p2, p3, layout=my.layout)  # Package::FunctionName
```


## Themes

A great deal of thought went into the default settings of ggplot2 to maximize the visual clarity of the graphs. However some people believe the defaults for many of the tiny graphical settings are poor. You can modify each of these but it is often easier to modify them all at once by selecting a different theme. The ggplot2 package includes several, `theme_bw()`, and `theme_minimal()` being the two that I use most often. Below are a few to examples:
```{r, fig.height=4}
Rmisc::multiplot( p1 + theme_bw(),          # Black and white
                  p1 + theme_minimal(),   
                  p1 + theme_dark(),        
                  p1 + theme_light(),
                  cols=2 )                         # two columns of graphs   
```

There are more themes in the package `ggthemes`
```{r, fig.height=4}
library(ggthemes)
Rmisc::multiplot( p1 + theme_stata(),          # Black and white
                  p1 + theme_economist(),   
                  p1 + theme_fivethirtyeight(),        
                  p1 + theme_excel(),
                  cols=2 )                         # two columns of graphs     
```

Almost everything you want to modify can be modified within the theme and you should check out the `ggplot2` documentation for more information and examples of how to modify different elements. [http://docs.ggplot2.org/current/theme.html]

## Exercises

1. We'll next make some density plots that relate several factors towards the birthweight of a child.
    a) Load the `MASS` library, which includes the dataset `birthwt` which contains information about 189 babies and their mothers.
    b) Add better labels to the `race` and `smoke` variables using the following:
        ```{r, warning=FALSE, message=FALSE}
        library(MASS)
        library(dplyr)
        birthwt <- birthwt %>% mutate(
          race  = factor(race,  labels=c('White','Black','Other')),
          smoke = factor(smoke, labels=c('No Smoke', 'Smoke')))
        ```
    c) Graph a histogram of the birthweights `bwt` using `ggplot(birthwt, aes(x=bwt)) + geom_histogram()`.
    d) Make separate graphs that denote whether a mother smoked during pregnancy using the `facet_grid()` command.
    e) Perhaps race matters in relation to smoking. Make our grid of graphs vary with smoking status changing vertically, and race changing horizontally (that is the formula in `facet_grid()` should have smoking be the y variable and race as the x).
    f) Remove `race` from the facet grid, (so go back to the graph you had in part d). I'd like to next add an estimated density line to the graphs, but to do that, I need to first change the y-axis to be density (instead of counts), which we do by using `aes(y=..density..)` in the `ggplot()` aesthetics command.
    g) Next we can add the estimated smooth density using the `geom_density()` command.
    h) To really make this look nice, lets change the fill color of the histograms to be something less dark, lets use `fill='cornsilk'` and `color='grey60'`. To play with different colors that have names, check out the following: [http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf].
    i) Change the order in which the histogram and the density line are added to the plot. Does it matter and which do you prefer?

2. Load the dataset `ChickWeight` and remind yourself what the data was using ?ChickWeight. Using `facet_wrap()`, produce a scatter plot of weight vs age for each chick. Use color to distinguish the four different `Diet` treatments.



