# Data Reshaping

```{r, echo=FALSE}
# Unattach any packages that happen to already be loaded. In general this is unecessary
# but is important for the creation of the book to not have package namespaces
# fighting unexpectedly.
pkgs = names(sessionInfo()$otherPkgs)
if( length(pkgs > 0)){
  pkgs = paste('package:', pkgs, sep = "")
  for( i in 1:length(pkgs)){
    detach(pkgs[i], character.only = TRUE, force=TRUE)
  }
}
```

```{r, warning=FALSE, message=FALSE}
# library(tidyr)   # for the gather/spread commands
# library(dplyr)   # for the join stuff
library(tidyverse) # dplyr, tidyr, ggplot2, etc.
```


Most of the time, our data is in the form of a data frame and we are interested in exploring the relationships. However most procedures in R expect the data to show up in a 'long' format where each row is an observation and each column is a covariate. In practice, the data is often not stored like that and the data comes to us with repeated observations included on a single row. This is often done as a memory saving technique or because there is some structure in the data that makes the 'wide' format attractive. As a result, we need a way to convert data from 'wide' to 'long' and vice-versa.

Next we need a way to squish two data frames together. It is often advantagous to store data that would be be repeated seperately in a different table so that a particular piece of information lives in only one location. This makes the data easier to modify, and more likely to maintain consistence. However, this practice requires that, when necessary, we can add information to a table, that might involve a lot of duplicated rows.

## `cbind` & `rbind`
Base R has two functions for squishing two data frames together, but they assume that the data frames are aligned correctly. The `c` and `r` parts of `cbind` and `rbind` correspond to if we are pushing columns together or rows together.

```{r}
# Define a tibble using a rowwise layout.
df1 <- tribble(            
  ~ID,  ~First,
  1,   'Alice',
  2,   'Bob',
  3,   'Charlie')

# Define a tibble by columns
df2 <- tibble( Last = c('Anderson', 'Barker', 'Cooper') )

# Squish the two tibbles together by columns
People <- cbind( df1, df2 )
People
```

Squishing together by rows is easy as well.
```{r}
df3 <- tibble( ID=4, First='Daniel', Last='Davidson' )

People <- rbind( People, df3 )
People
```


## `tidyr`

There is a common issue with obtaining data with many columns that you wish were organized as rows. For example, I might have data in a grade book that has several homework scores and I'd like to produce a nice graph that has assignment number on the x-axis and score on the y-axis. Unfortunately this is incredibly hard to do when the data is arranged in the following way:

```{r}
grade.book <- rbind(
  data.frame(name='Alison',  HW.1=8, HW.2=5, HW.3=8, HW.4=4),
  data.frame(name='Brandon', HW.1=5, HW.2=3, HW.3=6, HW.4=9),
  data.frame(name='Charles', HW.1=9, HW.2=7, HW.3=9, HW.4=10))
grade.book

```


What we want to do is turn this data frame from a *wide* data frame into a *long* data frame. In MS Excel this is called pivoting. Essentially I'd like to create a data frame with three columns: `name`, `assignment`, and `score`. That is to say that each homework datum really has three pieces of information: who it came from, which homework it was, and what the score was. It doesn't conceptually matter if I store it as 3 rows of 4 columns or 12 rows so long as there is a way to identify how a student scored on a particular homework. So we want to reshape the HW1 to HW4 columns into two columns (assignment and score). 

This package was built by the same people that created dplyr and ggplot2 and there is a nice introduction at: [http://blog.rstudio.org/2014/07/22/introducing-tidyr/]

### Verbs 
As with the dplyr package, there are two main verbs to remember:

1. `gather` - Gather multiple columns that are related into two columns that contain the original column name and the value. For example for columns HW1, HW2, HW3 we would gather them into two column HomeworkNumber and Score. In this case, we refer to HomeworkNumber as the key column and Score as the value column. So for any key:value pair you know everything you need.

2. `spread` - This is the opposite of gather. This takes a key column (or columns) and a results column and forms a new column for each level of the key column(s).

```{r}
# first we gather the score columns into columns we'll name Assesment and Score
tidy.scores <- grade.book %>% 
  gather( key=Homework,     # What should I call the key column
          value=Score,      # What should I call the values column
          HW.1:HW.4)        # which columns to apply this to
tidy.scores
```

To spread the key:value pairs out into a matrix, we use the `spread()` command. 

```{r}
# Turn the Assessment/Score pair of columns into one column per factor level of Assessment
tidy.scores %>% spread( key=Homework, value=Score )
```

One way to keep straight which is the `key` column is that the key is the category, while `value` is the numerical value or response. 

## Storing Data in Multiple Tables
In many datasets it is common to store data across multiple tables, usually with the goal of minimizing memory used as well as providing minimal duplication of information so any change that must be made is only made in a single place.

To see the rational why we might do this, consider building a data set of blood donations by a variety of donors across several years. For each blood donation, we will perform some assay and measure certain qualities about the blood and the patients health at the donation.

```{r, echo=FALSE}
donations <- data.frame( 
  Donor = c('Derek', 'Jeff'), 
  Hemoglobin=c(17.4, 16.9), 
  Systolic=c(121, 145), 
  Diastolic=c(80, 101) )
donations
```

But now we have to ask, what happens when we have a donor that has given blood multiple times?  In this case we should just have multiple rows per person along with a date column to uniquely identify a particular donation.

```{r, echo=FALSE}
donations <- data.frame( 
  Donor = c('Derek', 'Derek', 'Jeff'), 
  Date = lubridate::mdy( c('4-14-2017', '6-20-2017', '8-14-2017') ),
  Hemoglobin=c(17.4, 16.5, 16.9), 
  Systolic=c(120, 121, 145), 
  Diastolic=c(79, 80, 101) )
```
```{r}
donations
```

I would like to include additional information about the donor where that infomation doesn't change overtime. For example we might want to have information about the donar's birthdate, sex, blood type.  However, I don't want that information in _every single donation line_.  Otherwise if I mistype a birthday and have to correct it, I would have to correct it _everywhere_. For information about the donor, should live in a `donors` table, while information about a particular donation should live in the `donations` table.

Furthermore, there are many Jeffs and Dereks in the world and to maintain a unique identifier (without using Social Security numbers) I will just create a `Donor_ID` code that will uniquely identify a person.  Similarly I will create a `Donation_ID` that will uniquely identify a dontation.

```{r, echo=FALSE}
donations <- data.frame(
  Donation_ID = c('Donation_1', 'Donation_2', 'Donation_3'),
  Donor_ID = c('Donor_1', 'Donor_1', 'Donor_2'), 
  Date = lubridate::mdy( c('4-14-2017', '6-20-2017', '8-14-2017') ),
  Hemoglobin=c(17.4, 16.5, 16.9), 
  Systolic=c(120, 121, 145), 
  Diastolic=c(79, 80, 101) )

donors <- data.frame(
  Donor_ID = c('Donor_1', 'Donor_2'),
  F_Name = c('Derek', 'Jeff'),
  L_Name = c('Lee', 'Smith'),
  B_Type = c('O+', 'A'),
  Birth = lubridate::mdy( c('9-17-1976', '6-23-1974') ),
  Street = c('7392 Willard', '873 Vine'),
  City = c('Flagstaff', 'Bozeman'),
  State = c('AZ','MT') )
```
```{r}
donors
donations
```

If we have a new donor walk in and give blood, then we'll have to create a new entry in the `donors` table as well as a new entry in the `donations` table. If an experienced donor gives again, we just have to create a new entry in the donations table.

```{r, echo=FALSE}
donors <- rbind(donors, data.frame(
  Donor_ID='Donor_3', 
  F_Name='Aubrey', 
  L_Name='Lee', 
  B_Type='O+', 
  Birth = lubridate::mdy( c('12-15-1980') ),
  Street='7392 Willard', 
  City='Flagstaff', 
  State='AZ'))
donations <- rbind(donations, data.frame(
  Donation_ID = c('Donation_4', 'Donation_5'),
  Donor_ID = c('Donor_1','Donor_3'),
  Date = lubridate::mdy( c('8-26-2017', '8-26-2017') ),
  Hemoglobin=c(17.6, 16.1), 
  Systolic=c(120, 137), 
  Diastolic=c(79, 90) ))
```
```{r}
donors
donations
```


This data storage set-up might be flexible enough for us.  However what happens if somebody moves? If we don't want to keep the historical information, then we could just change the person's `Street_Address`, `City`, and `State` values.  If we do want to keep that, then we could create `donor_addresses` table that contains a `Start_Date` and `End_Date` that denotes the period of time that the address was valid.

```{r, echo=FALSE}
donor_addresses <- data.frame(
  Donor_ID = c('Donor_1', 'Donor_1', 'Donor_1', 'Donor_2', 'Donor_3'),
  Street = c('346 Treeline', '645 Main', '7392 Willard', '873 Vine', '7392 Willard'),
  City = c('Pullman', 'Flagstsff', 'Flagstaff', 'Bozeman', 'Flagstaff'),
  State = c('WA', 'AZ','AZ', 'MT', 'AZ'),
  Start_Date = lubridate::mdy( c('1-26-2015', '6-28-2016', '7-3-2017', '3-17-2015', '6-1-2017')),
  End_Date = lubridate::mdy( c('6-27-16', '7-2-2017', NA, NA, NA )) )
```
```{r}
donor_addresses
```

Given this data structure, we can now easily create new donations as well as store donor information. In the event that we need to change something about a donor, there is only _one_ place to make that change.

However, having data spread across multiple tables is challenging because I often want that information squished back together.  For example, the blood donations services might want to find all 'O' or 'O+' donors in Flagstaff and their current mailing address and send them some notification about blood supplies being low.  So we need someway to join the `donors` and `donor_addresses` tables together in a sensible manner.

## Table Joins
Often we need to squish together two data frames but they do not have the same number of rows. Consider the case where we have a data frame of observations of fish and a separate data frame that contains information about lake (perhaps surface area, max depth, pH, etc). I want to store them as two separate tables so that when I have to record a lake level observation, I only input it *one* place. This decreases the chance that I make a copy/paste error. 

To illustrate the different types of table joins, we'll consider two different tables.
```{r}
# tibbles are just data.frames that print a bit nicer and don't automatically
# convert character columns into factors.  They behave a bit more consistently
# in a wide variety of situations compared to data.frames.
Fish.Data <- tibble(
  Lake_ID = c('A','A','B','B','C','C'), 
  Fish.Weight=rnorm(6, mean=260, sd=25) ) # make up some data
Lake.Data <- tibble(
  Lake_ID = c(    'B','C','D'),   
  Lake_Name = c('Lake Elaine', 'Mormon Lake', 'Lake Mary'),   
  pH=c(6.5, 6.3, 6.1),
  area = c(40, 210, 240),
  avg_depth = c(8, 10, 38))
```

```{r}
Fish.Data
Lake.Data
```

Notice that each of these tables has a column labled `Lake_ID`. When we join these two tables, the row that describes lake `A` should be duplicated for each row in the `Fish.Data` that corresponds with fish caught from lake `A`.

```{r}
full_join(Fish.Data, Lake.Data)
```

Notice that because we didn't have any fish caught in lake `D` and we don't have any Lake information about lake `A`, when we join these two tables, we end up introducing missing observations into the resulting data frame.

The other types of joins govern the behavor or these missing data.

**`left_join(A, B)`** For each row in A, match with a row in B, but don't create any more rows than what was already in A.

**`inner_join(A,B)`** Only match row values where both data frames have a value.

```{r}
left_join(Fish.Data, Lake.Data)
```

```{r}
inner_join(Fish.Data, Lake.Data)
```

The above examples assumed that the column used to join the two tables was named the same in both tables.  This is good practice to try to do, but sometimes you have to work with data where that isn't the case.  In that situation you can use the `by=c("ColName.A"="ColName.B")` syntax where `ColName.A` represents the name of the column in the first data frame and `ColName.B` is the equivalent column in the second data frame.


Finally, the combination of `gather` and `join` allows me to do some very complex calculations across many columns of a data set.  For example, I might gather up a set of columns, calculate some summary statistics, and then join the result back to original data set.  

```{r}
grade.book %>%
  group_by(name) %>%
  gather( key=Homework, value=Score, HW.1:HW.4 ) %>%
  summarise( HW.avg = mean(Score) ) %>%
  left_join( grade.book, . )
```


## Exercises

1. Suppose we are given information about the maximum daily temperature from a weather station in Flagstaff, AZ. The file is available at the GitHub site that this book is hosted on.

    ```{r}
    FlagTemp <-  read.csv(
      'https://github.com/dereksonderegger/570L/raw/master/data-raw/FlagMaxTemp.csv',
       header=TRUE, sep=',')
    ```
    This file is in a wide format, where each row represents a month and the columns X1, X2, ..., X31 represent the day of the month the observation was made. 
    a. Convert data set to the long format where the data has only four columns: `Year`, `Month`, `Day`, `Tmax`.
    b. Calculate the average monthly maximum temperature for each Month in the dataset (So there will be 365 mean maximum temperatures). *You'll probably have some issues taking the mean because there are a number of values that are missing and by default R refuses to take means and sums when there is missing data. The argument `na.rm=TRUE` to `mean()` allows you to force R to remove the missing observations before calculating the mean. Alternatively at some point in your workflow, you could remove rows that include a missing values using `dplyr::drop_na`.* 
    c. Convert the average month maximums back to a wide data format where each line represents a year and there are 12 columns of temperature data (one for each month) along with a column for the year. *There will be a couple of months that still have missing data because the weather station was out of commision for those months and there was NO data for the entire month.*
    
2. A common task is to take a set of data that has multiple categorical variables and create a table of the number of cases for each combination. An introductory statistics textbook contains a dataset summarizing student surveys from several sections of an intro class. The two variables of interest for us are `Gender` and `Year` which are the students gender and year in college.
    a. Download the dataset and correctly order the `Year` variable using the following:
        ```{r, eval=FALSE}
        Survey <- read.csv('http://www.lock5stat.com/datasets/StudentSurvey.csv', na.strings=c('',' ')) 
        ```
    b. Using some combination of `dplyr` functions, produce a data set with eight rows that contains the number of responses for each gender:year combination. Make sure your table orders the `Year` variable in the correct order of `First Year`, `Sophmore`, `Junior`, and then `Senior`. *You might want to look at the following functions: `dplyr::count` and `dplyr::drop_na`.* 
    c. Using `tidyr` commands, produce a table of the number of responses in the following form:
    
        |   Gender    |  First Year  |  Sophmore  |  Junior   |  Senior   |
        |:-----------:|:------------:|:----------:|:---------:|:---------:|
        |  **Female** |              |            |           |           |  
        |  **Male**   |              |            |           |           | 
    

3. We often are given data in a table format that is easy for a human to parse, but annoying a program. In the following example we have [data](https://github.com/dereksonderegger/570L/raw/master/data-raw/US_Gov_Budget_1962_2020.xls) of US government expenditures from 1962 to 2020. (I downloaded this data from https://obamawhitehouse.archives.gov/omb/budget/Historicals (Table 3.2) on Sept 22, 2019.) Our goal is to end up with a data frame with columns for `Function`, `Subfunction`, `Year`, and `Amount`. We'll ignore the "On-budget" and "Off-budget" distinction.
    a) Download the data file, inspect it, and read in the data using the `readxl` package.
    b) Rename the `Function or subfunction` column to `Department`.
    b) Remove any row with Total, Subtotal, On-budget or Off-budget.
    c) Create a new column for ID_number and parse the Function column for it.
    d) If all (or just 2015?) the year values are missing, then the `Department` corresponds to `Function` name. Otherwise `Department` corresponds to the `Subfunction`. Perhaps the easiest way to create the `Function` and `Subfunction` columns is to write a `for` loop that steps through the data row-by-row. Alternatively, the `tidyr::fill` function might be useful.
    e) Reshape the data into four columns for Function, Subfunction, Year, and Amount.
    f) Replace any Amount value of `..........` with an NA.
    g) Make a bar chart that compares spending for National Defense, Health, Medicare, Income Security, and Social Security for the years 2001 through 2015. *Notice you'll have to sum up the subfunctions within each function.*


4. Data table joins are extremely common because effective database design almost always involves having multiple tables for different types of objects. To illustrate both the table joins and the usefulness of multiple tables we will develop a set of data frames that will represent a credit card company's customer data base. We will have tables for Customers, Retailers, Cards, and Transactions.  Below is code that will create and populate these tables.
    ```{r}
    Customers <- tribble(
      ~PersonID, ~Name, ~Street, ~City, ~State,
      1, 'Derek Sonderegger',  '231 River Run', 'Flagstaff', 'AZ',
      2, 'Aubrey Sonderegger', '231 River Run', 'Flagstaff', 'AZ',
      3, 'Robert Buscaglia', '754 Forest Heights', 'Flagstaff', 'AZ',
      4, 'Roy St Laurent', '845 Elk View', 'Flagstaff', 'AZ')
    
    Retailers <- tribble(
      ~RetailID, ~Name, ~Street, ~City, ~State,
      1, 'Kickstand Kafe', '719 N Humphreys St', 'Flagstaff', 'AZ',
      2, 'MartAnnes', '112 E Route 66', 'Flagstaff', 'AZ',
      3, 'REI', '323 S Windsor Ln', 'Flagstaff', 'AZ' )
    
    Cards <- tribble(
      ~CardID, ~PersonID, ~Issue_Date, ~Exp_Date,
      '9876768717278723',  1,  '2019-9-20', '2022-9-20',
      '5628927579821287',  2,  '2019-9-20', '2022-9-20',
      '7295825498122734',  3,  '2019-9-28', '2022-9-28',
      '8723768965231926',  4,  '2019-9-30', '2022-9-30' ) 
    
    Transactions <- tribble(
      ~CardID, ~RetailID, ~Date, ~Amount,
      '9876768717278723', 1, '2019-10-1', 5.68,
      '7295825498122734', 2, '2019-10-1', 25.67,
      '9876768717278723', 1, '2019-10-2', 5.68,
      '9876768717278723', 1, '2019-10-2', 9.23,
      '5628927579821287', 3, '2019-10-5', 68.54,
      '7295825498122734', 2, '2019-10-5', 31.84,
      '8723768965231926', 2, '2019-10-10', 42.83) 
    
    Cards <- Cards %>% 
      mutate( Issue_Date = lubridate::ymd(Issue_Date),
              Exp_Date   = lubridate::ymd(Exp_Date) )
    Transactions <- Transactions %>% 
      mutate( Date = lubridate::ymd(Date))
    ```
    a) Create a table that gives the credit card statent for Derek. It should give all the transactions, the amounts, and the store name. Write your code as if the only initial information you have is the customer's name.
    b) Aubrey has lost her credit card on Oct 15, 2019. Close her credit card and issue her a new credit card in the `Cards` table.
    c) Aubrey is using her card at Kickstand Kafe on Oct 16, 2019 for coffee with a charge of $4.98. Generate a new transaction for this action. 
    d) On Oct 17, 2019, some nefarious person is trying to use her credit card at REI. Make sure your code in part (c) first checks to see if the credit card is active before creating a new transaction. Using the same code, verify that this transaction is denied.
    e) Generate a table that gives the credit card statement for Aubrey. It should give all the transactions, amounts, and retailer name for both credit cards she had during this period.
    
  
5. The package `nycflights13` contains information about all the flights that arrived in or left from New York City in 2013. This package contains five data tables, but there are three data tables we will work with. The data table `flights` gives information about a particular flight, `airports` gives information about a particular airport, and `airlines` gives information about each airline. Create a table of all the flights on February 14th by Virgin America that has columns for the carrier, destination, departure time, and flight duration. Join this table with the airports information for the destination. Notice that because the column for the destination airport code doesn't match up between `flights` and `airports`, you'll have to use the `by=c("TableA.Col"="TableB.Col")` argument where you insert the correct names for `TableA.Col` and `TableB.Col`.


