---
output:
  pdf_document: default
  html_document: default
---
# Graphing Part II

```{r, echo=FALSE}
# Un-attach any packages that happen to already be loaded. In general this is unnecessary
# but is important for the creation of the book to not have package name spaces
# fighting unexpectedly.
pkgs = names(sessionInfo()$otherPkgs)
if( length(pkgs > 0)){
  pkgs = paste('package:', pkgs, sep = "")
  for( i in 1:length(pkgs)){
    detach(pkgs[i], character.only = TRUE, force=TRUE)
  }
}

knitr::opts_chunk$set(cache=TRUE, fig.height=4, fig.width=6)
```


```{r, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)   # loading ggplot2 and dplyr
library(viridis)     # The viridis color schemes
library(latex2exp)   # For plotting math notation

library(plotly)     # for interactive hover-text

library(maps)              # package with data about country borders
library(sf)                # Simple Features for GIS
library(rnaturalearth)     # package with detailed information about country &
library(rnaturalearthdata) # state/province borders, and geographical features
# devtools::install_github('ropensci/rnaturalearthhires')
library(rnaturalearthhires)
library(leaflet)
```

We have already seen how to create many basic graphs using the `ggplot2` package. However we haven't addressed many common scenarios. In this chapter we cover many graphing tasks that occur.

## Multi-plots
There are several cases where it is reasonable to need to take several possibly unrelated graphs and put them together into a single larger graph. This is not possible using `facet_wrap` or `facet_grid` as they are intended to make multiple highly related graphs. Instead we have to turn to other packages that enhance the `ggplot2` package.

### `cowplot` package
Claus O. Wilke wrote a lovely [book](https://serialmentor.com/dataviz/) about data visualization and also wrote an R package to help him tweek his plots. One of the functions in his `cowplot` package is called `plot_grid` and it takes in any number of plots and lays them out on a grid.

```{r, fig.height=4}
P1 <- ggplot(iris, aes(x=Sepal.Length, y=Sepal.Width, color=Species)) + 
  geom_point(size=3) + theme(legend.position='bottom')
P2 <- ggplot(trees, aes(x=Height, y=Volume, color=Girth)) + 
  geom_point() + theme(legend.position='bottom')
P3 <- ggplot(iris, aes(x=Sepal.Length)) + 
  geom_histogram(bins=30)
P4 <- ggplot(iris, aes(x=Species, y=Sepal.Length, fill=Species)) + 
  geom_boxplot() + theme(legend.position='bottom')

cowplot::plot_grid(P1, P2, P3, P4)
```

Notice that the graphs are by default are arranged in a 2x2 grid.  We could  adjust the number or rows/columns using the `nrow` and `ncol` arguments. Furthermore, we could add labels to each graph so that the figure caption to refer to "Panel A" or "Panel B" as appropriate using the `labels` option.

```{r, fig.height=3}
cowplot::plot_grid(P2, P3, P4, nrow=1, labels=c('A','B','C'))

```

### `multiplot` in `Rmisc`
In his book, *R Graphics Cookbook* Winston Chang produced a really nice function to address this issue, but just showed the code. The folks that maintain the R miscellaneous package `Rmisc` kindly included his function. The benefit of using this function is that you can control the layout to not be on a grid.  For example we might want two graphs side by side, and then the third be short and wide underneath both. By specifying different numbers and rows and columns in my layout matrix, we can highly customize how the plot looks.

```{r}
# Define where the first plot goes, etc.
my.layout <- matrix(c(1, 2, 2,
                      1, 2, 2,
                      3, 3, 3), byrow = TRUE, nrow=3)

Rmisc::multiplot(P2, P3, P4, layout = my.layout )
```

Unfortunately, `Rmisc::multiplot` doesn't have a label option, so if you want to refer to "Panel A", you need to insert the label into each plot separately.


## Customized Scales

While `ggplot` typically produces very reasonable choices for values o the axis scales and color choices for the `color` and `fill` options, we often want to tweak them.

### Color Scales


#### Manually Select Colors

For an individual graph, we might want to set the color manually.  Within `ggplot` there are a number of `scale_XXX_` functions where the `XXX` is either `color` or `fill`.

```{r}
cowplot::plot_grid(
  P1 + scale_color_manual( values=c('red','navy','forest green') ),
  P2 + scale_color_gradient(low = 'blue', high='red')
)
```

For continuous scales for fill and color, there is also a `scale_XXX_gradient2()` function which results in a *divergent* scale where you set the `low` and `high` values as well as the midpoint color and value. There is also a `scale_XXX_grandientn()` function that allows you to set as many colors as you like to move between.

```{r}
cowplot::plot_grid(
  P2 + scale_color_gradient2( low = 'black', mid='white', midpoint=14, high='red' ),
  P2 + scale_color_gradientn(colors = c('red','orange','yellow','green','blue','violet') )
)
```

Generally I find that I make poor choices when picking colors manually, but there are times that it is appropriate.


#### Palettes
In choosing color schemes, a good approach is to use a color palette that has already been created by folks that know about how colors are displayed and what sort of color blindness is possible. There are two palette options that we'll discuss, but there are a variety of other palettes available by downloading a package.

##### `RColorBrewer` palettes
Using the `ggplot::scale_XXX_brewer()` functions, we can easily work with the package `RColorBrewer` which provides a nice set of color palettes. These palettes are separated by purpose. 

**Qualitative** palettes employ different hues to create visual differences between classes. These palettes are suggested for nominal or categorical data sets.
```{r, echo=FALSE}
# Palettes for Qualitative Categories
RColorBrewer::display.brewer.all(type='qual')
```

**Sequential** palettes progress from light to dark. When used with interval data, light colors represent low data values and dark colors represent high data values.
```{r, echo=FALSE}
# Palettes for Ordered Categories
RColorBrewer::display.brewer.all(type='seq')
```

**Diverging** palettes are composed of darker colors of contrasting hues on the high and low extremes and lighter colors in the middle.
```{r, echo=FALSE}
# Ordered Categories with a neutral value in the middle
RColorBrewer::display.brewer.all(type='div')
```


To use one of these palettes, we just need to pass the palette name to `scale_color_brewer` or `scale_fill_brewer`
```{r}
cowplot::plot_grid(
  P1 + scale_color_brewer(palette='Dark2'),
  P4 + scale_fill_brewer(palette='Dark2') )
```


##### `viridis` palettes

The package `viridis` sets up a few different color palettes that have been well thought out and maintain contrast for people with a variety of color-blindess types as well as being converted to grey-scale.

```{r, echo=FALSE}
palettes <- c('viridis','magma','inferno','plasma','cividis')
Plots <- list()
for(i  in 1:length(palettes)){
  Plots[[i]] <- 
    data.frame(x=seq(0,1,by=0.001)) %>%
    ggplot( aes(x=x, y=1, fill=x)) +
    geom_raster() + 
    cowplot::theme_map() + 
    theme(legend.position = 'none') +
    labs(title=palettes[i]) +
    scale_fill_viridis(option=palettes[i])
}

Rmisc::multiplot(plotlist = Plots )
```


```{r}
cowplot::plot_grid(
  P1 + scale_color_viridis_d(option='plasma'),      # _d for discrete
  P2 + scale_color_viridis_c( option='viridis') )   # _c for continuous
```


There are a bunch of other packages that manage color palettes such as `paletteer`, `ggsci` and `wesanderson`. 


### Setting major & minor ticks
For continuous variables, we need to be able to control what tick and grid lines are displayed.  In `ggplot`, there are *major* and *minor* breaks and the major breaks are labeled and minor breaks are in-between the major breaks. The break point labels can also be set.

```{r}
ggplot(trees, aes(x=Height, y=Volume)) + geom_point() +
  scale_x_continuous( breaks=seq(65,90, by=5), minor_breaks=65:90 ) +
  scale_y_continuous( breaks=c(30,50), labels=c('small','big') )
```

### Log Scales
For this example, we'll use the `ACS` data from the `Lock5Data` package that has information about `Income` (in thousands of dollars) and `Age`. Lets make a scatter plot of the data.
```{r, fig.height=3}
# Import the data and drop any zeros
data('ACS', package='Lock5Data') 
ACS <- ACS %>%
  drop_na() %>% filter(Income > 0)

cowplot::plot_grid(
  ggplot(ACS, aes(x=Age, y=Income)) +
    geom_point(),
  ggplot(ACS, aes(x=Age, y=log10(Income))) +
    geom_point()
)
```

Plotting the raw data results in an ugly graph because six observations dominate the graph and the bulk of the data (income < $100,000) is squished together. One solution is to plot income on the $\log_{10}$ scale. The second graph does that, but the labeling is all done on the log-scale and most people have a hard time thinking in terms of logs.

This works quite well to see the trend of peak earning happening in a persons 40s and 50s, but the scale is difficult for me to understand (what does $\log_{10}\left(X\right)=1$ mean here? Oh right, that is $10^{1}=X$ so that is the $10,000 line). It would be really nice if we could do the transformation but have the labels on the original scale. 

```{r, fig.height=3}
cowplot::plot_grid(
  ggplot(ACS, aes(x=Age, y=Income)) +
    geom_point() +
    scale_y_log10(),
  ggplot(ACS, aes(x=Age, y=Income)) +
    geom_point() +
    scale_y_log10(breaks=c(1,10,100),
                  minor=c(1:10,
                        seq( 10, 100,by=10 ),
                        seq(100,1000,by=100))) +
    ylab('Income (1000s of dollars)')
)

```

Now the y-axis is in the original units (thousands of dollars) but obnoxiously we only have two labeled values. Lets define the major break points (the white lines that have numerical labels) to be at 1,10,100 thousand dollars in salary. Likewise we will tell `ggplot2` to set minor break points at 1 to 10 thousand dollars (with steps of 1 thousand dollars) and then 10 thousand to 100 thousand but with step sizes of 10 thousand, and finally minor breaks above 100 thousand being in steps of 100 thousand.


## Themes

Many fine-tuning settings in `ggplot2` can be manipulated using the `theme()` function. I've used it previously to move the legend position, but there are many other options. 
```{r}
p1 <- ggplot(ChickWeight, aes(x=Time, y=weight, colour=Diet, group=Chick)) +
    geom_line() + labs(title='Chick Weight: Birth to 21 days')

# Two common examples of things to change
cowplot::plot_grid(nrow=2,
  p1 + theme(plot.title = element_text(hjust = 0.5, size=25)),  
  p1 + theme(legend.position = 'bottom')                 # legend to bottom
)
```

There are many things to tweak using the `theme()` command and to get a better idea of what is possible, I recommend visiting the ggplot2 web page documentation and [examples](https://ggplot2.tidyverse.org/reference/theme.html#examples). Notably, one thing that is NOT changed using the `theme` command is the color scales.  


A great deal of thought went into the default settings of ggplot2 to maximize the visual clarity of the graphs. However some people believe the defaults for many of the tiny graphical settings are poor. You can modify each of these but it is often easier to modify them all at once by selecting a different theme. The ggplot2 package includes several, `theme_bw()`, and `theme_minimal()` being the two that I use most often. Other packages, such as `cowplot` and `ggthemes`, have a bunch of other themes that you can select from. Below are a few  examples:
```{r, fig.height=6}
Rmisc::multiplot( cols = 2,
  p1 + theme_bw()      + labs(title='theme_bw'), # Black and white
  p1 + theme_minimal() + labs(title='theme_minimal'),   
  p1 + theme_dark()    + labs(title='theme_dark'),        
  p1 + theme_light()   + labs(title='theme_light')
)
```

```{r, fig.height=8}
Rmisc::multiplot( cols = 2,
  p1 + cowplot::theme_cowplot() + labs(title='cowplot::theme_cowplot()'),
  p1 + cowplot::theme_minimal_grid() + labs(title='cowplot::theme_minimial_grid'),
  p1 + ggthemes::theme_stata() + labs(title='ggthemes::theme_stata()'),
  p1 + ggthemes::theme_tufte() + labs(title='ggthemes::theme_tufte()'),
  p1 + ggthemes::theme_economist() + labs(title='ggthemes::theme_economist()'),
  p1 + ggthemes::theme_fivethirtyeight() + labs(title='ggthemes::theme_fivethirtyeight()'),
  p1 + ggthemes::theme_excel_new() + labs(title='ggthemes::theme_excel_new()'),
  p1 + ggthemes::theme_gdocs() + labs(title='ggthemes::theme_gdocs()')
  )
```




Finally, we might want to select a theme for all subsequent plots or modify a specific aspect of the theme.

|   Command                  |     Result            |
|:--------------------------:|:----------------------|
| `theme_set( theme_bw() )`  | Set the default theme to be the `theme_bw()` theme. |
| `theme_update( ... )`      | Update the current default them.  |

This will allow you to set the graphing options at the start of your Rmarkdown/R-script document. However the one thing it does not do is allow you to change the default color themes (we still have to do for each graph).


## Mathematical Notation

It would be nice to be able to include mathematical formula and notation on plot axes, titles, and text annotation. R plotting has a notation scheme which it calls `expressions`. You can learn more about how R `expressions` are defined by looking at the [plotmath help](https://stat.ethz.ch/R-manual/R-devel/library/grDevices/html/plotmath.html) help page. They are similar to LaTeX but different enough that it can be frustrating to use. It is particularly difficult to mix character strings and math symbols. I recommend not bothering to learn R expressions, but instead learn LaTeX and use the R package `latex2exp` that converts character strings written in LaTeX to be converted into R's expressions.  

LaTeX is an extremely common typesetting program for mathematics and is widely used. The key idea is that `$` will open/close the LaTeX mode and within LaTeX mode, using the backslash represents that something special should be done.  For example, just typing `$alpha$` produces $alpha$, but putting a backslash in front means that we should interpret it as the greek letter alpha. So in LaTeX, `$\alpha$` is rendered as $\alpha$. We've already seen an [introduction](https://dereksonderegger.github.io/444/rmarkdown-tricks.html#mathematical-expressions) to LaTeX in the Rmarkdown Tricks chapter.

However, because I need to write character strings with LaTeX syntax, and R also uses the backslash to represent special characters, then to get the backslash into the character string, we actually need to do the same double backslash trick we did in the string manipulations using regular expressions section.

```{r}
seed <- 7867
N <- 20
data <- data.frame(x=runif(N, 1, 10)) %>%    # create a data set to work with
  mutate(y = 12 - 1*x + rnorm(N, sd=1))      # with columns x, y

model <- lm( y ~ x, data=data)               # Fit a regression model
data <- data %>%                             # save the regression line yhat points
  mutate(fit=fitted(model))

ggplot(data, aes(x=x)) +
  geom_point(aes(y=y)) +
  geom_line(aes(y=fit), color='blue') +
  annotate('text', x=9, y=9.5, 
           label=latex2exp::TeX('$\\hat{\\rho}$ = 0.916') ) +   # always double \\
  labs( x=latex2exp::TeX('X-axis $\\alpha$'), 
        y=latex2exp::TeX('Y: $\\log_{10}$(Income)'),
        title=latex2exp::TeX('Linear Models use: $\\hat{\\beta} = (X^TX)^{-1}X^Ty$'))
```

The warning message that is produced is coming from `ggplot` and I haven't figured out how to avoid it. Because it is giving us the graph we want, I'm just going to ignore the error for now.

One issue is how to add expression to a data frame. Unfortunately, neither `data.frame` nor `tibble` will allow a column of expressions, so instead we have store it as a character string. Below, we create three character strings using standard LaTeX syntax, and then convert it to a character string that represents the R expression. Finally, in `ggplot`, we tell the `geom_text` layer to parse the label and interpret it as an R expression.

```{r}
foo <- data.frame( x=c(1,2,3), y=c(2,2,2) ) %>%
  mutate( label1 = paste('$\\alpha$ = ', x) ) %>%       # label is a TeX character string
  mutate( label2 = latex2exp::TeX(label1, output = 'character') )  # label2 is an expression character string

ggplot(foo, aes(x=x, y=y) ) +
  geom_label( aes(label=label2), parse=TRUE )   # parse=TRUE forces an expression interpretation 
```

## Interactive plots with `plotly`

Plotly is technical computing company that specializes in data visualization. They have created an open source JavaScript library to produce graphs, which is confusingly referred to as plotly. Because plotly is JavaScript, and RStudio's Viewer pane is actually a web browser, it easily provides interactive abilities in RStudios Viewer pane. 

A good tutorial book about using plotly was written by [Carson Sievert](https://plotly-r.com).

The simple version is that you can take a ggplot graph and pipe it into the `ggplotly` and it will be rendered into an interactive version of the graph.

```{r, warning=FALSE, message=FALSE, eval=FALSE}
P1 <- trees %>%
  ggplot(aes(y=Volume, x=Height)) +
  geom_point()

P1 %>% plotly::ggplotly()
```

```{r, eval=FALSE, echo=FALSE}
# Previously I signed up for an account with plotly and 
# then I got an API key for uploading things to the server.
# 
# Inside my home directory, I added a .Rprofile file and 
# in there I have my user name and API key.
# 
# Sys.setenv("plotly_username"="zdereksonderegger")
# Sys.setenv("plotly_api_key"="DQ4Cmsdjklewrkje3pkdhYM") # this isn't correct
#   
P1 %>% 
  plotly::ggplotly() %>%
  api_create(filename='444_Plotly_Graph_2') 

# The api_create command saves the plot to the
# plotly server so it works in the online notes.
# This is not necessary for local Rstudio work.
# From here, instead of uploading the graph every
# time I re-knit the notes, I'm just linking to the 
# plot.
```

<iframe width="900" height="400" frameborder="0" scrolling="no" src="//plot.ly/~zdereksonderegger/3.embed"></iframe>

We can use the widgets to zoom in and out of the graph. Notice that as we hover over the point, it tells us the x/y coordinates. To add information to the hover-text, we just need to add a `text` aesthetic mapping to the plot. 

```{r, warning=FALSE, message=FALSE, eval=FALSE}
P1 <- trees %>%
  mutate(Obs_ID = row_number()) %>%  
  ggplot(aes(y=Volume, x=Height, 
             text=paste('Girth: ', Girth,   '\n',          # add some extra text
                        'Obs #: ', Obs_ID,  sep=''))) +    # to the hover information
  geom_point() 

P1 %>% ggplotly() 
```

```{r, warning=FALSE, message=FALSE, eval=FALSE, echo=FALSE}
P1 <- trees %>%
  mutate(Obs_ID = row_number()) %>%  
  ggplot(aes(y=Volume, x=Height, 
             text=paste('Girth: ', Girth,   '\n',          # add some extra text
                        'Obs #: ', Obs_ID,  sep=''))) +    # to the hover information
  geom_point() 

P1 %>% ggplotly() %>%
  api_create(filename='444_Plotly_Graph_3',
             overwrite = TRUE)              # Save to plotly server so it
                                            # it works in the online notes.
                                            # Not necessary for local Rstudio
                                            # work.
```


<iframe width="900" height="400" frameborder="0" scrolling="no" src="//plot.ly/~zdereksonderegger/5.embed"></iframe>


## Geographic Maps

We often need to graph countries or U.S. States. We might then fill the color of the state or countries by some variable. To do this, we need information about the shape and location of each country within some geographic coordinate system. The easiest system to work from is Latitude (how far north or south of the equator) and Longitude (how far east or west the prime meridian). 

### Package `maps`
The R package `maps` is one of the easiest way to draw a country or state map because it is built into the `ggplot` package. Unfortunately it is fairly US specific.

Because we might be interested in continents, countries, states/provinces, or counties, in the following discussion we'll refer to the geographic area of interest as a *region*. For `ggplot2` to interact with GIS type objects, we need a way to convert a GIS database of regions into a `data.frame` of a bunch of data points about the region's borders, where each data point is a Lat/Long coordinate and the region and sub-region identifiers. Then, to produce a map, we just draw a path through the data points. For regions like Hawaii's, which are composed of several non-contiguous areas, we include sub-regions so that the boundary lines don't jump from island to island.

Once we have the `data.frame` of regions that we are interested in selected, all we need to do is draw polygons in `ggplot2`.
```{r, fig.height=4, fig.width=6}
# ggplot2 function to create a data.frame with world level information
geo.data <- ggplot2::map_data('world') # Using maps::world database. 

# group: which set of points are contiguous and should be connected
# order: what order should the dots be connected
# region: The name of the region of interest
# subregion: If there are sub-regions with greater region
head(geo.data)

# Now draw a nice world map, 
ggplot(geo.data, aes(x = long, y = lat, group = group)) +
  geom_polygon( colour = "white", fill='grey50') 
```

The `maps` package has several data bases of geographical regions.  

|  Database    |  Description                               |
|:------------:|:-------------------------------------------|
| `world`      |  Country borders across the globe          |
| `usa`        |  The country boundary of the United States |
| `state`      |  The state boundaries of the United States |
| `county`     |  The county boundaries within states of the United States |
| `lakes`      |  Large fresh water lakes across the world  |
| `italy`      |  Provinces in Italy                        |
| `france`     |  Provinces in France                       |
| `nz`         |  North and South Islands of New Zealand    |

From within each of these databases, we can select to just return a particular region. So for example, we can get all the information we have about Ghana using the following:
```{r, fig.height=3, fig.width=3}
ggplot2::map_data('world', regions='ghana') %>%
  ggplot( aes(x=long, y=lat, group=group)) +
  geom_polygon( color = 'white', fill='grey40')
```

The `maps` package also has a `data.frame` of major US cities.  

```{r, fig.height=5, fig.width=5}
az.cities <- maps::us.cities %>%              # Lat/Long of major US cities
  filter(country.etc == 'AZ') %>%             # Only the Arizona Cities
  mutate(name = str_remove(name, '\\sAZ') )   # remove ' AZ' from the city name
  
small.az.cities <- az.cities %>%
  filter(name %in% c('Flagstaff','Prescott','Lake Havasu City','Yuma','Tucson',
                     'Sierra Vista','Phoenix','Tempe','Scottsdale'))

ggplot2::map_data('state', regions='arizona') %>%
  ggplot( aes(x=long, y=lat)) +
  geom_polygon( aes(group=group), color = 'white', fill='grey40') +
  geom_point(data=az.cities) +
  ggrepel::geom_label_repel(data=small.az.cities, aes(label=name))
```


### Packages `sf` and `rnaturalearth`

The `maps` package is fairly primitive in the data it has as well as the manner in which it stores the data. Another alternative is to use the *spatial features* package `sf` along with an on-line data base of GIS information from [Natural Earth](https://www.naturalearthdata.com).

There is a nice set of tutorials for the `sf` package.
[Part 1](https://www.r-spatial.org/r/2018/10/25/ggplot2-sf.html),
[Part 2](https://www.r-spatial.org/r/2018/10/25/ggplot2-sf-2.html), and
[Part 3](https://www.r-spatial.org/r/2018/10/25/ggplot2-sf-3.html).

From the Natural Earth package, we can more easily obtain information about world countries and state/provinces. There is also information about urban areas and roads as well as geographical details such as rivers and lakes.

```{r}
ne_countries(continent='Africa', returnclass = 'sf') %>%  # grab country borders in Africa
  ggplot() + geom_sf() +
  labs(title='Africa')
```

```{r}
ne_states(country='Ghana', returnclass = 'sf') %>% # grab provinces within Ghana
  ggplot() +
  geom_sf( ) + 
  labs(title='Ghana')
```

```{r}
# The st_centroid function takes the sf object and returns the center of 
# the polygon, again as a sf object.
Ghana <- ne_states(country='Ghana', returnclass = 'sf')  # grab provinces within Ghana
Ghana %>% ggplot() +
  geom_sf( ) + 
  geom_text( data=st_centroid(Ghana), size=2,
             aes(x=longitude, y=latitude, label=woe_name)) +
  labs(title='Ghana Administrative Regions')
```


### Package `leaflet`

Leaflet is a popular open-source JavaScript library for interactive maps. The package `leaflet` provides a nice interface to this package. The [tutorial](https://rstudio.github.io/leaflet/) for this package is quite good.

The basic work flow is:

1. Create a map widget by calling leaflet().
2. Create and add layers (i.e., features) to the map by using layer functions 
    a) `addTiles`
    b) `addMarkers` 
    c) `addPolygons`
3. Repeat step 2 as desired.
4. Print the map widget to display it.

```{r}
map <- leaflet() %>%  # Build a base map
  addTiles()  %>%     # Add the default tiles 
  addMarkers(lng=-1*(111+37/60+52/3600), lat=35+11/60+57/3600, popup="Flagstaff, AZ")
map %>% print()
```

Because we have added only one marker, then leaflet has decided to zoom in as much as possible. If we had multiple markers, it would have scaled the map to include all of them.

As an example of an alternative, I've downloaded a GIS shape file of forest service administrative area boundaries.

```{r}
# The shape file that I downloaded had the CRS format messed up. I need to 
# indicate the projection so that leaflet doesn't complain.
Forest_Service <- 
  sf::st_read('data-raw/Forest_Service_Boundaries/S_USA.AdministrativeRegion.shp') %>%
  sf::st_transform('+proj=longlat +datum=WGS84')

leaflet() %>%
  addTiles() %>%
  addPolygons(data = Forest_Service) %>%
  setView(-93, 42, zoom=3)
```



## Exercises

1. [Gapminder](https://www.gapminder.org) is a organization devoted to educating people about world facts and dispelling misconceptions between countries. 
    a) From their [data](https://www.gapminder.org/data/) page, download information about CO2 emissions per person. Import the data and filter out all but the latest year.
    b) Create a map of CO2 emissions per person . 

2. The `infmort` data set from the package `faraway` gives the infant mortality rate for a variety of countries. The information is relatively out of date, but will be fun to graph. T
    a) The `rownames()` of the dataset give the country names and you should create a new column that contains the country names. 
    b) Save the `ggplot2::map_data('world')` result as `geo.data`. Compare the `infmort` country names to the `region` names in the `geo.data` data set and do whatever is necessary to get the country names to concur. *Hint: `unique(x)` will return the unique items in a vector, `intersect(x,y)`returns the elements common to `x` and `y`, and `setdiff(x,y)`  returns the elements of `x` that are not in `y`. I recommend creating a `Country_Dictionary` that contains the standardized country name and all the weird versions you've run into. Every time you see some new abbreviation, just add it to the dictionary. To start with, you might consider all the ways people commonly refer to the United States of America.*
        ```{r}
        Country_Dictionary <- tribble(
          ~raw, ~standardized,
          'U.S.A',         'United States of America',
          'United States', 'United States of America',
          'US',            'United States of America')
        ```
        *After doing some simple data cleaning on the country names (e.g. swapping `_` for a space or vice versa), figure out which country names in the two data sets don't match up and decide on a standardized name and insert the translation into your dictionary table. Then join the dictionary to the `geo.data` and create a standardized country name column. Do the same for the `infmort` data and then the standardized country name levels should match up.*
    c) Join the `geo.data` with the `infmort` data.
    d) Make a map of the world where we shade in countries based on the country income. Set the color fill scale to be anything other than the default.
    e) Make a map of the world where we shade in the countries based on if they are oil exports. Color the map black if the country is an oil exporter, and a light gray if it is not.

3. The `infmort` data set can also be visualized using by faceting using `region` and `oil` export status. 
    a) Create scatter plots the countries income and infant mortality using a $log_{10}$ transformation for both axes.
    b) The package `ggrepel` contains functions `geom_text_repel()` and `geom_label_repel()` that mimic the basic `geom_text()` and `geom_label()` functions in `ggplot2`, but work to make sure the labels don't overlap. Select 10-15 countries to label and do so using the `geom_text_repel()` function.

4. Using the `datasets::trees` data, complete the following:
    a) Create a regression model for Volume as a function of Height.
    b) Using the `summary` command, get the y-intercept and slope of the regression line.
    c) Using `ggplot2`, graph the data.
    d) Create a nice white filled rectangle to add text information to using by adding the following annotation layer.
        ```{r, eval=FALSE}
          + annotate('rect', xmin=7, xmax=14.5, ymin=64, ymax=70,
                   fill='white', color='black')
        ```
    e) Add some annotation text to write the equation of the line $\hat{y}_i = -36.94 + 5.1 * x_i$ in the text area.
    f) Add annotation to add $R^2 = 0.934$
    
5. Using the `datasets::Titanic` data set, create a bar graph showing the number of individuals that survived or not. Make sure to include the passenger `Class`, `Sex`, and `Age` variable information. *Unfortunately, the data is stored as a `table` and to expand it to a data frame, the following code can be used.*
    ```{r}
    Titanic <- Titanic %>% as.data.frame()
    ```
    a) Make this graph using the default theme.
    b) Make this graph using the `theme_bw()` theme.
    c) Make this graph using the `cowplot::theme_minimal_hgrid()` theme.
    d) Why would it be beneficial to drop the vertical grid lines?
    