# Graphing Maps

```{r, echo=FALSE}
# Un-attach any packages that happen to already be loaded. In general this is unnecessary
# but is important for the creation of the book to not have package namespaces
# fighting unexpectedly.
pkgs = names(sessionInfo()$otherPkgs)
if( length(pkgs > 0)){
  pkgs = paste('package:', pkgs, sep = "")
  for( i in 1:length(pkgs)){
    detach(pkgs[i], character.only = TRUE, force=TRUE)
  }
}
knitr::opts_chunk$set(cache=FALSE)
```


```{r, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)   # loading ggplot2 and dplyr

library(sf)                # Simple Features for GIS

library(rnaturalearth)     # package with detailed information about country &
library(rnaturalearthdata) # state/province borders, and geographical features
# devtools::install_github('ropensci/rnaturalearthhires')
library(rnaturalearthhires) # Hi-Resolution Natural Earth

# devtools::install_github("UrbanInstitute/urbnthemes")
# devtools::install_github("UrbanInstitute/urbnmapr")
library(urbnmapr)
#library(urbnthemes)

library(leaflet)
```

## Introduction{-}

We often have data that is associated with some sort of geographic information. 
For example, we might have information based on US state counties. It would be 
nice to be able to produce a graph where we fill in the county with a color shade
associated with our data (This is called a chloropleth).  
Or perhaps put dots at the center of each county and the dots might be color
coded or size coded to our data. 
But the critical aspect is that we already have some data, but we need some
additional data relating to the shape of the county or state of interest.

There are several tutorials that I've found useful while preparing this introduction:

*   There is a simple 
    [blog style series of posts](https://www.r-spatial.org/r/2018/10/25/ggplot2-sf.html) 
    that is a quick read. 

*   The `sf` package vignettes are a really great
    [resource](https://r-spatial.github.io/sf/reference/sf.html).

*   Finally there is a great on-line 
    [book](https://keen-swartz-3146c4.netlify.com) 
    by Edzer Pebesma and Roger Bivand about mapping in R. 



### Coordinate Reference Systems (CRS){-}

There are many ways to represent a geo-location. 

1. **WGS84 aka Latitude/Longitude** One of the oldest systems, and most well known,
is the latitude/longitude grid. The latitude measures north/south where zero is the
equator and +90 and -90 are the north and south poles. Longitude is the east/west
measurement and the zero is the 
[Prime Meridian](https://en.wikipedia.org/wiki/Prime_meridian) 
(which is close to the Greenwich Meridian) and +180 and -180 are the anti-meridian
near the Alaska/Russia border. The United States is in the negative longitudes. 
The problem with latitude/longitude is that small
differences in lat/long coordinates near the equator are large distance, but near
the poles it would be much much smaller. Another weirdness is that lat/long 
coordinates are often given in a base 60 system of degrees/minutes/seconds.  
To get the decimal version we use the formula
$$\textrm{Decimal Value} = \textrm{Degree} + \frac{\textrm{Minutes}}{60} + \frac{\textrm{Seconds}}{3600}$$
For example, Flagstaff AZ has lat/long 35°11′57″N 111°37′52″W.  Notice that the
longitude is 111 degrees W which should actually be negative.  This gives a lat/long
pair of:
    ```{r}
    35 + 11/60 + 57/3600
    -1 * (111 + 37/60 + 52/3600)
    ```

2. **Reference Point Systems** A better idea is to establish a grid of reference
points along the grid of the planet and then use offsets from those reference 
points. One of the most common projection system is the 
[Universal Transverse Mercator (UTM)](https://en.wikipedia.org/wiki/Universal_Transverse_Mercator_coordinate_system) 
which uses a grid of 60 reference points. From these reference points, we will 
denote a location using a northing/easting offsets. The critical concept is that
if you are given a set of northing/easting coordinates, we also need the reference
point and projection system information. For simplicity we'll refer to both the
lat/long or northing/easting as the coordinates. 


### Projections
No matter how the geographic information is stored, we need to decide how to
display the. For small geographic areas, this choice isn't too important, but for
large areas it is. The fundamental reason is because the earth is approximately
a sphere and any 2-dimensional map must distort the size and/or shape of the areas
being represented.

To tell R which projection to use, we'll modify the CRC to include information
about what project we want. Inside a geographic object, all the information about
the CRC and projection is encoded in a single text string and we just need to
update that.

```{r}
world_map <- 
  ne_countries(returnclass = 'sf') %>%  # tibble containing country shapes
  ggplot() + 
  geom_sf(fill='grey') 
world_map
```

The default projection is the 'Patterson' projection and gives far to much pixel
area to areas in the arctic when looking at the entire planet. Alternatives include
the Robinson and Mollweide.

```{r}
P1 <- world_map +
  coord_sf(crs = '+proj=moll') +      # change the projection to Mollweide
  labs(title='Mollweide Projection')
P2 <- world_map +
  coord_sf(crs = '+proj=robin') +    # change the projection to Robinson
  labs(title='Robinson Projection')

cowplot::plot_grid(P1, P2)
```

When graphing smaller regions, the default projection doesn't quite work because
Alaska spans across the international date line. 

```{r}
USA_map <- ne_states(country = 'united states of america', returnclass = 'sf') %>%
  ggplot() +
  geom_sf(fill='grey') 
USA_map
```

A somewhat better alternative when showing just the contiguous United States is 
to use the UTM projection which includes a `zone`
[option](https://earth-info.nga.mil/GandG/coordsys/grids/utm.html)
and center the map on the center zone.

```{r}
USA_map +
  coord_sf(crs = '+proj=utm +zone=12')     # change the projection to UTM zone 12
```

But Alaska and Hawaii are making this map a bit ugly and we often see them just
moved and re-sized. The `urbnmapr` package has data sets of the US states and
counties that already has a good projection that already does this.
```{r}
# pull the CRS information from Urban Institutes maps.
get_urbn_map(map='states', sf=TRUE) %>%
  ggplot() +
  geom_sf(fill='grey') +
  theme_void()  # get rid of latitude and longitude now I've moved things around
```


### Spatial Objects{-}
Regardless of the coordinate reference system (CRS) used, there are three major types of data that we might want to store. 

1.  **Points** The simplest type of data object to store is a single location.
    Storing them simply requires knowing the coordinates and the reference system.
    It is simple to keep track of a number of points as you just have a data frame
    of coordinates and then the reference system.

2.  **LineString** This maps out a one-dimensional line across the surface of the
    globe. An obvious example is to define a road.  To do this, we just need to
    define a sequence of points (where the order matters) and the object implicitly
    assumes the points are connected with straight lines. To represent an arbitrary
    curve, we simply need to connect points that are quite close together. As always,
    this object also needs to keep track of the CRS.

3.  **Polygons** These are similar to a LineString in that they are a sequential
    vector of points, but now we interpret them as forming an enclosed area so
    the line starts and ends at the same place. As always, we need to keep the CRS
    information for the points. We also will allow "holes" in the area by having
    one or more interior polygons cut out of it. To do this, we need to keep a
    list of removed polygons.

Each of the above type of spatial objects can be grouped together, much as we
naturally made a group of points. For example, the object that defines the borders
of the United States of America needs to have multiple polygons because each contiguous
land mass needs its own polygon. For example the state of Hawaii has 8 main islands and
129 minor islands. Each of those needs to be its own polygon for our "connect the dots"
representation to make sense.

Until recently, the spatial data structures in R (e.g. the `sp` package) required
users to keep track of the data object type as well as the reference system. This
resulted in code that contained many mysterious input strings that were almost
always poorly understood by the user. Fortunately there is now a formal ISO standard
that describes how spatial objects should be represented digitally and the types of
manipulations that users should be able to do. The R package `sf`, which stands for
"Simple Features" implements this standard and is quickly becoming the preferred R
library for handling spatial data.


### Tiles{-}
Instead of building a map layer by layer, you might want to start with some base
level information, perhaps some topological map with country and state names along
with major metropolitan areas. Tiles provide a way to get all the background map
information for you to then add your data on top. 

Tiles come in two flavors. Rasters are similar to a pictures in that every pixel
is stored. These have the downside that when zooming in, the pixelation becomes
obvious. Vector based tiles actually only store the underlying spatial information
and handle zooming in without pixelation. 

## Obtaining Spatial Data{-}

Often I already have some information associated with some geo-political unit such
as state or country level rates of something (e.g. country's average life span, or
literacy rate). Given the country name, we want to produce a map with the data we
have encoded with colored dots centered on the country or perhaps fill in the 
country with shading associated with the statistic of interest. To do this, we
need data about the shape of each country! 

In general it is a bad idea to rely on spatial data that is static on a user's
machine. First, large scale geo-political borders, coastal boundaries can 
potentially change. Second, fine scale details like roads and building locations
are constantly changing. Third, the level of detail needed is quite broad for 
world maps, but quite small for neighborhood maps. It would be a bad idea to 
download neighborhood level data for the entire world in the chance the a use 
might want fine scale detail for a particular neighborhood. However, it is also 
a bad idea to query a web-service every time I knit a document together. Therefore 
we will consider ways to obtain the information needed for a particular scale 
and save it locally but our work flow will always include a data refresh option.


### Natural Earth Database{-}

[Natural Earth](https://www.naturalearthdata.com) is a public domain map database
and is free for any use, both commercial and non-commercial. There is a nice R 
package, `rnaturalearth` that provides convenient interface. There is also 
information about urban areas and roads as well as geographical details such 
as rivers and lakes. There is a mechanism to download data at different 
resolutions as well as matching functions for reading in the data from a local 
copy of it.

There are a number of data sets that are automatically downloaded with the
`rnaturalearth` package including country and state/province boarders.

```{r}
ne_countries(continent='Africa', returnclass = 'sf') %>%  # grab country borders in Africa
  ggplot() + geom_sf() +
  labs(title='Africa')
```

```{r}
ne_states(country='Ghana', returnclass = 'sf') %>% # grab provinces within Ghana
  ggplot() +
  geom_sf( ) + 
  labs(title='Ghana')
```

```{r}
# The st_centroid function takes the sf object and returns the center of 
# the polygon, again as a sf object.
Ghana <- ne_states(country='Ghana', returnclass = 'sf')  # grab provinces within Ghana
Ghana %>% ggplot() +
  geom_sf( ) + 
  geom_text( data=st_centroid(Ghana), size=2,
             aes(x=longitude, y=latitude, label=woe_name)) +
  labs(title='Ghana Administrative Regions')
```


There is plenty of other geographic information that you can download from 
Natural Earth. In the table below, scale refers to how large the file is and 
so scale might more correctly be interpreted as the data resolution.

| category   |  type   |  scale `small`  |  scale `medium` | scale `large`   |
|:----------:|:-------:|:-------------:|:-------------:|:-------------:|
| `physical`  | `coastline` |          Yes | Yes | Yes |
| `physical`  | `land` |               Yes | Yes | Yes |
| `physical`  | `ocean` |              Yes | Yes | Yes |
| `physical`  | `lakes` |              Yes | Yes | Yes |
| `physical`  | `geographic_lines` |   Yes | Yes | Yes |
| `physical`  | `minor_islands` |      No  | No  | Yes |
| `physical`  | `reefs` |              No  | No  | Yes |
| `cultural` | `populated_places` |   Yes | Yes | Yes |
| `cultural` | `urban_areas`      |   No  | Yes | Yes |
| `cultural` | `roads`            |   No  | No  | Yes |


### Urban Institute
#### Package `urbnmapr`
This is a 
[package](https://urban-institute.medium.com/how-to-create-state-and-county-maps-easily-in-r-part-2-d21d57310c50)
developed by the Urban Institute and has two convenient data sets for US
states and counties. This is quite US-centric, but it works for me.

```{r}

AZ_map_df <- get_urbn_map(map='counties', sf=TRUE) %>%
  filter(state_name == 'Arizona')
str(AZ_map_df)
```

Notice that there is a FIPS code, which is the Federal Information Processing Standards
which includes a standard code for identifying geographic information. These are
slowly being phased out in favor of GNIS Feature ID, and INCITS standards but
FIPS is still widespread.

```{r, fig.height=5, fig.width=5, warning=FALSE, message=FALSE}
# I downloaded some information about Education levels from the American Community Survey 
# website  https://www.census.gov/acs/www/data/data-tables-and-tools/
# by selecting the Education topic and then used the filter option to select the state
# and counties that I was interested in. Both the 1 year and 5 year estimates 
#  didn't include the smaller counties (too much uncertainty).
# I then had reshape the data a bit to the following
# format. I ignored the margin-of-error columns and didn't worry about the 
# the Race, Hispanic, or Gender differences.

AZ_Education <- 
  read_csv('data-raw/AZ_Population_25+_BS_or_Higher.csv') %>%
  arrange(County) %>%
  rename(Percent_BS = 'Percent_BS+') %>%
  mutate(County = str_to_title(County),
         County = str_c(County , ' County'))

# Show the County percent of 25 or older population with BS or higher 
AZ_Education
```

Finally we can joint the geographic data set and the Education data and plot it
```{r}
AZ_map_df %>%
  left_join(AZ_Education, by=c('county_name' = 'County')) %>%
  ggplot() +
  geom_sf(aes(fill=Percent_BS)) +
  coord_sf(crs = '+proj=utm +zone=12 +ellps=WGS84 +towgs84=0,0,0')
  # coord_sf( crs = '+proj=natearth +lon_0=0 +datum=WGS84 +units=m +no_defs')
```



### Other non-sf packages
#### Package `maps`{-}
The R package `maps` is one of the easiest way to draw a country or state maps
because it is built into the `ggplot` package. This is one of the easiest ways
I know of to get US county information. Unfortunately it is fairly US specific.

Once we have the `data.frame` of regions that we are interested in selected,
all we need to do is draw polygons in `ggplot2`.
```{r, fig.height=4, fig.width=6}
# ggplot2 function to create a data.frame with world level information
geo.data <- ggplot2::map_data('world') # Using maps::world database. 

# group: which set of points are contiguous and should be connected
# order: what order should the dots be connected
# region: The name of the region of interest
# subregion: If there are sub-regions with greater region
head(geo.data)

# Now draw a nice world map, not using Simple Features,
# but just playing connect the dots.
ggplot(geo.data, aes(x = long, y = lat, group = group)) +
  geom_polygon( colour = "white", fill='grey50') 
```

The `maps` package has several data bases of geographical regions.  

|  Database    |  Description                               |
|:------------:|:-------------------------------------------|
| `world`      |  Country borders across the globe          |
| `usa`        |  The country boundary of the United States |
| `state`      |  The state boundaries of the United States |
| `county`     |  The county boundaries within states of the United States |
| `lakes`      |  Large fresh water lakes across the world  |
| `italy`      |  Provinces in Italy                        |
| `france`     |  Provinces in France                       |
| `nz`         |  North and South Islands of New Zealand    |


The `maps` package also has a `data.frame` of major US cities. This is a
nice place to get city locations, though I often want to turn this into
a `sf` data set with the following:

```{r}
Cities <- maps::us.cities %>%
  mutate(city = str_sub(name, 1, -3),    # remove the state Abbreviation
         city = str_trim(city)) %>%
  rename(state = 'country.etc') %>%
  select(city, state, lat, long) %>%
  st_as_sf(coords=c('long','lat'),           # Which columns have longitute/latitude
           crs="+proj=longlat +datum=WGS84") # I'm using long/lat coordinates
```


```{r}
AZ_Cities <- Cities %>% filter(state == 'AZ')
AZ_map_df %>%
  ggplot() +
  geom_sf(fill='grey') +
  geom_sf(data=AZ_Cities) +
  coord_sf(crs = '+proj=utm +zone=12 +ellps=WGS84 +towgs84=0,0,0')
```


Adding labels to the cities is fairly easy as well. We'll use the `ggrepel`
package to push the labels away from their data points but we need to specify to
use the `sf` geometry to get the x and y locations.

```{r}
AZ_Cities <- Cities %>% filter(state == 'AZ')
AZ_Cities_ToLabel <- AZ_Cities %>%
  filter(city %in% c('Phoenix','Tempe','Scottsdale',
                     'Flagstaff','Prescott','Lake Havasu City',
                     'Yuma','Tucson','Sierra Vista'))

AZ_map_df %>%
  ggplot() +
  geom_sf(fill='grey') +
  geom_sf(data=AZ_Cities) +
  ggrepel::geom_text_repel(
    data = AZ_Cities_ToLabel,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates") +
  coord_sf(crs = '+proj=utm +zone=12 +ellps=WGS84 +towgs84=0,0,0')
```




### Package `leaflet`{-}

Leaflet is a popular open-source JavaScript library for interactive maps. The 
package `leaflet` provides a nice interface to this package. The
[tutorial](https://rstudio.github.io/leaflet/) 
for this package is quite good.

The basic work flow is:

1. Create a map widget by calling leaflet().
2. Create and add layers (i.e., features) to the map by using layer functions 
    a)  `addTiles` - These are the background of the map that we will put stuff
        on top of.
    b)  `addMarkers` 
    c)  `addPolygons`
3. Repeat step 2 as desired.
4. Print the map widget to display it.

```{r, eval=knitr::is_html_output() }
map <- leaflet() %>%  # Build a base map
  addTiles()  %>%     # Add the default tiles 
  addMarkers(lng=-1*(111+37/60+52/3600), 
             lat=35+11/60+57/3600, 
             popup="Flagstaff, AZ")
map %>% print()
```

Because we have added only one marker, then leaflet has decided to zoom in as much
as possible. If we had multiple markers, it would have scaled the map to include
all of them.

As an example of an alternative, I've downloaded a GIS shape file of forest
service administrative area boundaries.

```{r, eval=knitr::is_html_output()}
# The shape file that I downloaded had the CRS format messed up. I need to 
# indicate the projection so that leaflet doesn't complain.
Forest_Service <- 
  sf::st_read('data-raw/Forest_Service_Boundaries/S_USA.AdministrativeRegion.shp') %>%
  sf::st_transform('+proj=longlat +datum=WGS84')

leaflet() %>%
  addTiles() %>%
  addPolygons(data = Forest_Service) %>%
  setView(-93, 42, zoom=3)
```

## Exercises
1.  [Gapminder](https://www.gapminder.org) is a organization devoted to educating
    people about world facts and dispelling misconceptions between countries
    a)  From their [data](https://www.gapminder.org/data/) page, download
        information about country per capita CO2 emissions. Import the data and
        filter out all but the latest year.
    b)  Create a `Country_Dictionary` that will map countries in the CO2 data to
        the regions in the `ggplot2::map_data('world')` table. In this problem we
        have two data sets we need to join, by country and region, but
        unfortunately those don’t always match.
        
        For example, in the CO2 data, there is one line for "Antigua and Barbuda”,
        but in the geo.data there is one region for Antigua and one for Barbuda. A
        clever solution is to make two rows in your country dictionary that
        matches both the regions “Antiqua" and "Barbuda" to the CO2 value for
        "Antigua and Barbuda”. This way when you join the CO2 data to your
        country_dictionary, the result is to copy the CO2 result from the combined
        “Antigua and Barbuda” to 1 row for Antigua and 1 row for Barbuda.
        ```{r, eval=FALSE}
            # You'll need to fill in the rest of the countries and regions that don't match!
            Country_Dictionary <- tribble(
              ~raw,                             ~standardized,
              "Antigua and Barbuda",            "Antigua",
              "Antigua and Barbuda",            "Barbuda",
              "Congo, Dem. Rep.",               "Democratic Republic of the Congo",
              "Congo, Rep.",                    "Democratic Republic of the Congo")
        ```
        For the rest you should be able to figure out the matching country except
        for perhaps that Cote d’Ivoire is the Ivory Coast and Swaziland is now
        Eswatini. For the Democratic Republic of the Congo and Republic of the
        Congo, we have to combine their CO2 values because the map data just has
        the DRC, so for any of the standardized counties with multiple rows, take
        the max CO2 value after we've joined the `CO2` data with the
        `Country_Dictionary`. *Hint: `unique(x)` will return the unique items in a
        vector, `intersect(x,y)`returns the elements common to `x` and `y`, and
        `setdiff(x,y)`  returns the elements of `x` that are not in `y`. Similarly
        the function `semi_join(A,B)` and `anti_join(A,B)` return the rows of A
        the will be matched with B or the rows of A the DON'T match with B.
    
    c)  Create a map of CO2 emissions per person. Select a color scale that
        emphasizes the countries with high per capita CO2 levels.
