# Databases

```{r, echo=FALSE}
# Un-attach any packages that happen to already be loaded. In general this is unecessary
# but is important for the creation of the book to not have package namespaces
# fighting unexpectedly.
pkgs = names(sessionInfo()$otherPkgs)
if( length(pkgs > 0)){
  pkgs = paste('package:', pkgs, sep = "")
  for( i in 1:length(pkgs)){
    detach(pkgs[i], character.only = TRUE, force=TRUE)
  }
}
```

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(DBI)         # DataBase Interface Package
library(dbplyr)      # dplyr with databases!
```

There is a YouTube [Video Lecture](https://youtu.be/ElDmEwslJw0) for the chapter.

As our data grows larger and is being updated more frequently, we need to stop using static input files and instead learn to interact with databases. There are a many reasons for using a database, but these are my favorite:

## Introduction

1. Data Freshness. Because the database holds the definitive copy of the data, there isn't a problem of using a .csv file that is months (or years) old. That means my results are constantly being updated with new data.
2. No Local Storage. Because the data lives on the database, I don't have to occupy gigabytes of space on my laptop to hold an out-of-date copy of the data.
3. Database actions are atomic. Whenever I update the database, the action either happens or it doesn't and the database should never be left in an inconsistent state. This extremely important when processing financial transactions, for example.

Fortunately, reading information from a database instead of an in-memory table won't change our current work flow and superficially the change is trivial. However, the impact can be quite profound in the timeliness and re-usability of our work.

<!-- Local databases are also quite useful.  -->

The great people at Rstudio have created a [great website](https://db.rstudio.com/overview/) for using databases using their `dbplyr` package.

However, the package `dbplyr` is really only intended for *reading* from the data base and does not support *writing* to the data base. 

## Establishing a Connection

Databases should be run on a server that is ALWAYS on and available via an internet connection. To connect to a database, we'll need to know the internet address and be able to authenticate with a username/password combination.

To demonstrate how a database works, we unfortunately need to have a live database to connect to. In real situations this would already be done (probably by somebody else) and you would just need to install some DataBase Interface (DBI) package that will negotiate creating a connection between your R-session and the database.

However for this example, we need to start up a data base before we can start working with.

```{r, eval=FALSE}
# Normally, a database connection looks something like this:
con <- DBI::dbConnect(RMariaDB::MariaDB(), 
  host = "database.rstudio.com",
  user = "hadley",
  password = rstudioapi::askForPassword("Database password")
)

# For a Postgres database, it might look like this:
con <- DBI::dbConnect(dbDriver("PostgresSQL"), dbname = "postgres",
  host = "database.nau.edu", port = 5432,
  user = "dls354", 
  password = rstudioapi::askForPassword("Database password"))
```

```{r}
# For our little toy example, we'll use a database I'll create right now.
# Establish a connection
con <- DBI::dbConnect(RSQLite::SQLite(), dbname = ":memory:")
```

Our final step involves populating our new database with some data so that we can play with it. For this example, we'll go back to the credit card data example from the data reshaping chapter. I have it available on the book's GitHub data-raw directory.
```{r, message=FALSE}
# Read in some data for the Credit Card example
stem <- 'https://raw.githubusercontent.com/dereksonderegger/444/master/data-raw/CreditCard'
stem <- '~/GitHub/444/data-raw/CreditCard'

Cards        <- read_csv(paste(stem, 'Cards.csv',        sep='_'), 
                         col_types = cols(CardID=col_character(),
                                          PersonID=col_character()))
Customers    <- read_csv(paste(stem, 'Customers.csv',    sep='_'),
                         col_types = cols(PersonID=col_character()))
Retailers    <- read_csv(paste(stem, 'Retailers.csv',    sep='_'),
                         col_types = cols(RetailID=col_character()))
Transactions <- read_csv(paste(stem, 'Transactions.csv', sep='_'),
                         col_types = cols(CardID=col_character(),
                                          RetailID=col_character()))

# Because the EXTREMELY simple RSQLite database doesn't support dates natively,
# we need to convert all the Date/Time values to be character strings. Notice
# because they are sorted as Year-Month-Day Hour:Minute:Second, sorting will
# still work.
# convert the Dates to character strings 
Cards <- Cards %>% 
  mutate(Issue_DateTime = as.character(Issue_DateTime),
         Exp_DateTime   = as.character(Exp_DateTime))
Transactions <- Transactions %>%
  mutate(DateTime = as.character(DateTime))
```

So that we remember what the data we are working with looks like:
```{r}
head(Customers)    # Key  is  PersonID
head(Cards)        # Keys are PersonID and CardID  
head(Transactions) # Keys are CardID and RetailID
head(Retailers)    # Key  is  RetailID
```

Critically, using the ID columns, we can take an individual transaction figure out what customer executed it. Finally I'll take these tables and load them into my RSQLite database.

```{r}
# Copy the tables to our newly set up database. The dbWriteTable() function is intended
# for database examples and is NOT how you would in practice create a database.
DBI::dbWriteTable(con, 'Cards', Cards,
                  field.types=c(CardID='character',PersonID='character',
                                Issue_DateTime='time',Exp_DateTime='time') )
DBI::dbWriteTable(con, 'Customers', Customers,
                  field.types=c(PersonID='character'))
DBI::dbWriteTable(con, 'Retailers', Retailers,
                  field.types=c(RetailID='character'))
DBI::dbWriteTable(con, 'Transactions', Transactions,
                  field.types=c(CardID='character', RetailID='character',
                                DateTime='time'))

rm(Cards, Customers, Retailers, Transactions)     # Remove all the setup except `con`
```


## Basic SQL Commands

The traditional way to interact with a database is by using SQL syntax. SQL stands for Structured Query Language and some understanding of SQL is mandatory for anyone that interacts with databases.  There are many good introduction to SQL but we'll cover a few basics here.

### SELECT rows

One way to run SQL commands on the remote database is to use the the `DBI` package commands.
There are several functions (`DBI::dbXXXXX()`)that will take a previously established connection and a SQL
command that is stored as a character string.

|  DBI Command   |  Description                    |
|:--------------:|:--------------------------------|
| `dbListTables()` | Returns the tables available in a database. |
| `dbListFields()` | What columns are in a given table. |
| `dbGetQuery()` | Run a SELECT command and return the result. |
| `dbColumnInfo()` | Returns meta information about columns returned by a `dbGetQuery()` command. |
| `dbExecute()`       | Run a SQL command that does not return a table (e.g. INSERT or UPDATE). The return value of this is the number of rows that were affected. |


For example, we might want to select all the rows and columns from the `Transactions` table. 
To do this, we would run the following SQL command.
```{r}
sql_cmd <- 'SELECT * FROM Transactions'        # Define a character string with SQL command
transactions <- DBI::dbGetQuery(con, sql_cmd)  # Run the command
transactions
```

We can examine the SQL command as follows:

|  SQL Function    |  Description                                |
|:----------------:|:--------------------------------------------|
| `SELECT`         |  A keyword that denotes that the following is a *query*. |
|  `*`             | A placeholder meaning all columns. This could be any column name(s). |
| `FROM`           | A keyword indicating that whatever follows is the table (or tables) being selected from. Any table joins need to be constrained in the WHERE clause to tell us what columns need to match.  |
| `WHERE`          | A keyword indicating the following logical statements will be used to filter rows. Boolean operators `AND`, `OR`, and `NOT` can be used to create complex filter statements. |


Because Rstudio supports code chunks with different languages (including SQL), you could  insert a SQL code chunk and have it appropriately run when knitting the document together. To run this interactively, you are required to have the output inline option set. (Preferences -> Rmarkdown -> Show output inline...)

A `sql` chunk requires the `connection` to be defined and optionally a output
variable name (`output.var`) to store the return value from the SQL call.

```
```{sql, connection=con, output.var='sql_output'}
/* This is a SQL code chunk! */
SELECT * from Customers
```
```
```{sql, connection=con, output.var='sql_output'}
/* This is a SQL code chunk! */
SELECT * from Customers;
```


```{r}
# In further R chunks, I can refer to the output.var variable
sql_output
```

*From here on out, I'll just note when I'm in an SQL chunk with a comment.*

SQL SELECT statement often include table joins. Unfortunately you have to always 
specify how to do the table joins and SQL won't default to using column names that
are common to both tables. In the following code we use `Table.Column` notation
to uniquely specify a table/column pair. Because SQL uses the dot in this manner,
you should avoid using a dot in your R column names.

In the following SQL command, the information about how to do the table joins
is mixed up with the information about filtering for a particular customer. I find
this more difficult to read.

```{sql, connection=con}
/* SQL Chunk */
SELECT Customers.Name, Transactions.DateTime, Retailers.Name, Transactions.Amount
  FROM Customers, Cards, Transactions, Retailers
  WHERE Customers.PersonID    = Cards.PersonID       AND 
        Cards.CardID          = Transactions.CardID  AND
        Transactions.RetailID = Retailers.RetailID   AND
        Customers.Name = 'Derek Sonderegger'
```



### INSERT a new Row
To insert a row into a table, we need to specify the table to insert into as well as
the values to be inserted. It is good practice to specify the column/value pair (as opposed
to assuming the column order) using the following syntax.
```{sql, connection=con}
/* SQL Chunk */
INSERT INTO Customers (PersonID, Name, Street, City, State)
VALUES (5, 'John Smith', '2378 N Main St', 'Flagstaff', 'AZ');
```


### UPDATE a row
Updating a row in the database is similar, but we need to specify which rows to 
update. This is done by including a "WHERE" clause, which feels similar to the 
`dplyr::filter()` function. The only major difference is that SQL uses the single `=`
sign for both assignment and testing equality.

```{sql, connection=con}
/* SQL Chunk */
UPDATE Customers
SET Street = '5638 Daisy Ln', City='Mountainaire', State='AZ'
WHERE PersonID = 3 AND Name='Robert Buscaglia';
```

In the WHERE statement, multiple test conditions must be separated by AND or OR conjunctions.

```{sql, connection=con}
/* SQL Chunk */
SELECT * from Customers
```


## Passing R variables into SQL chunks

Inevitably I would like to be able to create a SQL command that depends on some
dynamic R variable. For example, perhaps I'd like to write a script that queries
the user for a person's name and then prints out all the transactions associated 
with the person. To do this, we need a mechanism to pass an R variable into the
SQL command. 

```{r}
# In an R chunk, I've defined some object that I'd like to use in a SQL statement.
WhichRetailer <- 2
```

```{sql connection=con}
/* Anything with a ? prepended will be                              *
 * replaced with the R object of the same name.                     */
SELECT * FROM Transactions WHERE RetailID = ?WhichRetailer
```


There are some additional situations where a simple character string substitution
doesn't quite work because of how R interprets character strings.

<!-- A quick example of how the `glue` package is intended to work would be helpful here. -->
<!-- The package `glue` is intended make it easy to combine R variables with character -->
<!-- strings. -->

For example, selecting a person by name requires the character string to be quoted, and
some SQL databases have custom quotation syntax. So we'll use the `glue` package to
decide what the appropriate quotation syntax is. The curly brackets tell
`glue` that we want to work with R variable `customers` not the literal string.


```{r}
# R chunk
customer <- 'Derek Sonderegger'                      # For some databases, this works
customer <- glue::glue_sql('{customer}', .con=con)   # This should always work
customer
```

```{sql connection=con}
/* SQL Chunk */
SELECT * FROM Customers WHERE name IN (?customer)
```

If you want to have multiple variables in a SQL statement, we need to tell `glue_sql()`
to make a single SQL command using a `{r_vector*}` notation. The curly brackets tell
`glue` that we want to work with R variable `customers` not the literal string.

```{r}
# R chunk
customers <- c('Derek Sonderegger', 'Aubrey Sonderegger')
glue::glue_sql("{customers}", .con=con)
```
and if I append a `*`, that tells `glue` to make a single SQL statement, not two separate statements.
```{r}
# R Chunk
glue::glue_sql("{customers*}", .con=con)

```

Now we can happily select multiple people from our Customers table.
```{r}
# R Chunk
customer = glue::glue_sql("{customers*}", .con=con)
```

```{sql connection=con}
/* SQL Chunk */ 
SELECT * FROM Customers WHERE name IN (?customer)
```


These SQL statements can be whatever you'd like, and we can happily insert rows
into tables as well.
```{r}
# R Chunk
newCustomer <- c(4, 'Mike Wazowski', '1102 Main St, Apt A113', 'Phoenix', 'AZ')
newCustomer <- glue::glue_sql('{newCustomer*}', .con = con)
```

```{sql connection=con}
/* SQL Chunk */
INSERT INTO Customers (PersonID, Name, Street, City, State)
VALUES(?newCustomer)
```

And this has happily inserted Mike into our `Customers` table.
```{sql, connection=con}
/* SQL Chunk */
SELECT * FROM Customers
```







## Reading tables using `dbplyr`

There are a lot of good things about SQL, but for database queries, I would really like to pretend that the tables are in memory and use all of my favorite `dplyr` tools and pipelines. This would mean that I don't have to remember all the weird SQL syntax. However, the database interface `dbplyr` is ONLY intended for queries and NOT for updating or inserting rows into the tables.

The way this will work is that we will use the previously established database connection `con` to create a virtual link between the database table and some appropriately named R object.

```{r}
# connect the database tables to similarly named objects in R
Cards <-        tbl(con, 'Cards')
Customers <-    tbl(con, 'Customers')
Retailers <-    tbl(con, 'Retailers')
Transactions <- tbl(con, 'Transactions')
```

However, this does NOT download the whole table into R.  Instead it grabs only a couple of rows so that we can see the format. Notice that we don't know how many rows are in the Transactions table.
```{r}
Transactions %>% head(3)
# Transactions %>% tail(3)  # not supported because we haven't yet downloaded much information.
```

The guiding principle of `dbplyr` is to delay as much work for as long as possible actually pulling the data from the database. The rational is that we spend a great deal of time figuring out what the query should look like and too often we write a query that accidentally downloads millions of lines of data and slows down our network connection. Instead `dbplyr` returns just the first few rows of whatever query we are working on until we finish the pipeline with a `collect()` command that will cause us to download ALL of the query results and save them as a local `data.frame`.

```{r, warning=FALSE, message=FALSE}
CC_statement <- 
  Customers %>% 
  filter(Name == 'Derek Sonderegger') %>% select(PersonID) %>%
  left_join(Cards) %>% left_join(Transactions) %>% left_join(Retailers) %>%
  select(DateTime, Name, Amount) %>%
  rename(Retailer = Name) 

CC_statement
```

At this point, we *still* haven't downloaded all of the rows. Instead this is still a *lazy* query. To actually download everything, we'll pipe this into the `collect` function.

```{r}
CC_statement %>%
  collect()
```


It can be fun to see what the SQL code that is being generated is.


```{r}
CC_statement %>% show_query()
```

The algorithm used to convert my `dplyr` statement into a SQL statement doesn't mind nesting SQL statements and isn't the same as what I generated by hand, but it works.


The last step of a script should be to close the database connection.
```{r}
# Close our database connection when we are through...
dbDisconnect(con)
```

## Exercises {#Exercises_DataBases}
1. In this exercise, you'll create a database containing the `nycflights13` data. Make sure that you've already downloaded the `nycflights13` package.
    a. Create a SQLite database and connect to it using the following code: 
    ```{r, eval=FALSE}
    library(dplyr)
    # Start up a SQL-Lite database with the NYCFlights13 data pre-loaded
    con <- nycflights13_sqlite( )
    ```
    b. Through the `con` connection object, create links to the `flights` and `airlines` tables.
    c. From the `flights` table, summarize the percent of flights with a departure delayed by more than 10 minutes for each airline. Produce a table that gives the airline name (not the abbreviation) and the percent of flights that are late.
    d. Using the `dbDisconnect()` command to close the connection `con`. *If this throws an error, you might need to update your `dbplyr` package. The error for disconnecting the SQLite `nycflights13` database was fixed recently.*
    
2. For this exercise, we'll start a SQLite database and see that the SQLite application stores the data in a very specialized file structure, which usually has a file extension of `.db` or `.sqlite`.
    a. Create the SQLite database file in your current working directory using the following:
    ```{r, eval=FALSE}
    con <- DBI::dbConnect(RSQLite::SQLite(), dbname = "TestSQLiteFile.db")
    
    # Create a table using the iris data
    dbWriteTable(con, 'IRIS', iris)
    
    dbListTables(con)         # What tables are available...
    dbReadTable(con, "IRIS")  # Spit out the IRIS table
    dbDisconnect(con)         # Close connection
    ```
    b) Now check the files in your current working directory as there should now be a `TestSQLiteFile.db`. The SQLite file structure for data is extremely stable and works across platform types (Unix/Windows, 32/64 bit, big/little endian, etc).  As such, it is a good file type choice for storing lots of data in a compact format across different systems (e.g. applications that work on a mobile device vs a computer). While you can open this file using a text editor, you will only see the table declaration of column names and types. The data rows that follow will not be readable. 

