# Databases

```{r, echo=FALSE}
# Un-attach any packages that happen to already be loaded. In general this is unecessary
# but is important for the creation of the book to not have package namespaces
# fighting unexpectedly.
pkgs = names(sessionInfo()$otherPkgs)
if( length(pkgs > 0)){
  pkgs = paste('package:', pkgs, sep = "")
  for( i in 1:length(pkgs)){
    detach(pkgs[i], character.only = TRUE, force=TRUE)
  }
}
```

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(DBI)         # DataBase Interface Package
library(dbplyr)      # dplyr with databases!
```

There is a YouTube [Video Lecture](https://youtu.be/ElDmEwslJw0) for the chapter.

## Introduction

As our data grows larger and is being updated more frequently, we need to stop 
using static input files and instead learn to interact with databases. There are 
a many reasons for using a database, but these are my favorite:

1.  **Data Freshness** Because the database holds the definitive copy of the data, 
    there isn't a problem of using a .csv file that is months (or years) old. That 
    means my results are constantly being updated with new data. When working with
    a team of individuals or applications that all have read/write access to the
    data, there must be some program making sure data is not accidentally overwritten
    or out of date.

2.  **Non-Local Storage** Often we deal with databases that live on a computer that
    we access via the internet. Because the data lives on the database and not my
    computer, I don't have to  occupy gigabytes of space on my laptop to hold an 
    out-of-date copy of the data. Furthermore, this allows us to interact with datasets
    that are too large to even be held on my wimpy laptop.

3.  **Application Reliability** Whenever I update the database, the action either 
    happens or it doesn't and the database should never be left in an inconsistent 
    state. As a result, financial transactions can be reliably processed over
    unreliable connections and the worst that happens is that the transaction 
    doesn't happen. For applications where the database is also on the same 
    device, this reliability means that if the application (game!) is abruptly
    ended, the game state isn't corrupted and we don't lose the user's whole
    history.

Fortunately, reading data with databases isn't too hard. If all we need to do is
read table and calculate summary statistics, the `dbplyr` package allows us to 
almost ignore that that data isn't locally stored. The great people at Rstudio 
have created a 
[great website](https://db.rstudio.com/overview/) 
for reading data from databases using their `dbplyr` package.

For updating a database, we'll need to learn a little SQL and learn how to pass 
R variables into the SQL call and import the SQL result back into our R environment.


## Establishing a Connection

Databases should be run on a server that is ALWAYS on and available via an 
internet connection. To connect to a database, we'll need to know the internet 
address and be able to authenticate with a username/password combination.

To demonstrate how a database works, we unfortunately need to have a live 
database to connect to. In real situations this would already be done (probably 
by somebody else) and you would just need to install some DataBase Interface (DBI) 
package that will negotiate creating a connection between your R-session and the 
database.

```{r, eval=FALSE}
# Normally, a database connection looks something like this:
con <- DBI::dbConnect(RMariaDB::MariaDB(), 
  host = "database.rstudio.com",
  user = "hadley",
  password = rstudioapi::askForPassword("Database password")
)

# For a Postgres database, it might look like this:
con <- DBI::dbConnect(dbDriver("PostgresSQL"), dbname = "postgres",
  host = "database.nau.edu", port = 5432,
  user = "dls354", 
  password = rstudioapi::askForPassword("Database password"))
```

However for this tutorial, we need to start up a data base before we can start 
working with.

```{r}
# For our little toy example, we'll use a database I'll create right now.
# Establish a connection
con <- DBI::dbConnect(RSQLite::SQLite(), dbname = ":memory:")
```

Our final step involves populating our new database with some data so that we 
can play with it. For this example, we'll go back to the credit card data example 
from the data reshaping chapter. I have it available on the book's GitHub data-raw 
directory.
```{r, message=FALSE}
# Read in some data for the Credit Card example
stem <- 'https://raw.githubusercontent.com/dereksonderegger/444/master/data-raw/CreditCard'
stem <- '~/GitHub/444/data-raw/CreditCard'

Cards        <- read_csv(paste(stem, 'Cards.csv',        sep='_'), 
                         col_types = cols(CardID=col_character(),
                                          PersonID=col_character()))
Customers    <- read_csv(paste(stem, 'Customers.csv',    sep='_'),
                         col_types = cols(PersonID=col_character()))
Retailers    <- read_csv(paste(stem, 'Retailers.csv',    sep='_'),
                         col_types = cols(RetailID=col_character()))
Transactions <- read_csv(paste(stem, 'Transactions.csv', sep='_'),
                         col_types = cols(CardID=col_character(),
                                          RetailID=col_character()))

# Because the EXTREMELY simple RSQLite database doesn't support dates natively,
# we need to convert all the Date/Time values to be character strings. Notice
# because they are sorted as Year-Month-Day Hour:Minute:Second, sorting will
# still work.
# convert the Dates to character strings 
Cards <- Cards %>% 
  mutate(Issue_DateTime = as.character(Issue_DateTime),
         Exp_DateTime   = as.character(Exp_DateTime))
Transactions <- Transactions %>%
  mutate(DateTime = as.character(DateTime))
```

So that we remember what the data we are working with looks like:
```{r}
head(Customers)    # Key  is  PersonID
head(Cards)        # Keys are PersonID and CardID  
head(Transactions) # Keys are CardID and RetailID
head(Retailers)    # Key  is  RetailID
```

Critically, using the ID columns, we can take an individual transaction figure 
out what customer executed it. Finally I'll take these tables and load them into 
my RSQLite database.

```{r}
# Copy the tables to our newly set up database. The dbWriteTable() function is intended
# for database examples and is NOT how you would in practice create a database.
DBI::dbWriteTable(con, 'Cards', Cards,
                  field.types=c(CardID='character',PersonID='character',
                                Issue_DateTime='time',Exp_DateTime='time') )
DBI::dbWriteTable(con, 'Customers', Customers,
                  field.types=c(PersonID='character'))
DBI::dbWriteTable(con, 'Retailers', Retailers,
                  field.types=c(RetailID='character'))
DBI::dbWriteTable(con, 'Transactions', Transactions,
                  field.types=c(CardID='character', RetailID='character',
                                DateTime='time'))

rm(Cards, Customers, Retailers, Transactions)     # Remove all the setup except `con`
```


## Basic SQL Commands

The traditional way to interact with a database is by using SQL syntax. SQL stands 
for Structured Query Language and some understanding of SQL is mandatory for anyone 
that interacts with databases.  There are many good introduction to SQL but we'll 
cover a few basics here.

### SELECT rows

One way to run SQL commands on the remote database is to use the the `DBI` package 
commands.
There are several functions (`DBI::dbXXXXX()`)that will take a previously established 
connection and a SQL command that is stored as a character string.

|  DBI Command   |  Description                    |
|:--------------:|:--------------------------------|
| `dbListTables()` | Returns the tables available in a database. |
| `dbListFields()` | What columns are in a given table. |
| `dbGetQuery()` | Run a SELECT command and return the result. |
| `dbColumnInfo()` | Returns meta information about columns returned by a `dbGetQuery()` command. |
| `dbExecute()`       | Run a SQL command that does not return a table (e.g. INSERT or UPDATE). The return value of this is the number of rows that were affected. |


For example, we might want to select all the rows and columns from the `Transactions` table. 
To do this, we would run the following SQL command.
```{r}
sql_cmd <- 'SELECT * FROM Transactions'        # Define a character string with SQL command
transactions <- DBI::dbGetQuery(con, sql_cmd)  # Run the command
transactions
```

We can examine the SQL command as follows:

|  SQL Function    |  Description                                |
|:----------------:|:--------------------------------------------|
| `SELECT`         |  A keyword that denotes that the following is a *query*. |
|  `*`             | A placeholder meaning all columns. This could be any column name(s). |
| `FROM`           | A keyword indicating that whatever follows is the table (or tables) being selected from. Any table joins need to be constrained in the WHERE clause to tell us what columns need to match.  |
| `WHERE`          | A keyword indicating the following logical statements will be used to filter rows. Boolean operators `AND`, `OR`, and `NOT` can be used to create complex filter statements. |


Because Rstudio supports code chunks with different languages (including SQL), 
you could  insert a SQL code chunk and have it appropriately run when knitting 
the document together. To run this interactively, you are required to have the 
output inline option set. (Preferences -> Rmarkdown -> Show output inline...)

A `sql` chunk requires the `connection` to be defined and optionally a output
variable name (`output.var`) to store the return value from the SQL call.

```
```{sql, connection=con, output.var='sql_output'}
/* This is a SQL code chunk! */
SELECT * from Customers
```
```
```{sql, connection=con, output.var='sql_output'}
/* This is a SQL code chunk! */
SELECT * from Customers;
```


```{r}
# In further R chunks, I can refer to the output.var variable
sql_output
```

*From here on out, I'll just note when I'm in an SQL chunk with a comment.*

SQL SELECT statement often include table joins. Unfortunately you have to always 
specify how to do the table joins and SQL won't default to using column names that
are common to both tables. In the following code we use `Table.Column` notation
to uniquely specify a table/column pair. Because SQL uses the dot in this manner,
you should avoid using a dot in your R column names.

In the following SQL command, the information about how to do the table joins
is mixed up with the information about filtering for a particular customer. I find
this more difficult to read.

```{sql, connection=con}
/* SQL Chunk */
SELECT Customers.Name, Transactions.DateTime, Retailers.Name, Transactions.Amount
  FROM Customers, Cards, Transactions, Retailers
  WHERE Customers.PersonID    = Cards.PersonID       AND 
        Cards.CardID          = Transactions.CardID  AND
        Transactions.RetailID = Retailers.RetailID   AND
        Customers.Name = 'Derek Sonderegger'
```



### INSERT a new Row
To insert a row into a table, we need to specify the table to insert into as well as
the values to be inserted. It is good practice to specify the column/value pair (as opposed
to assuming the column order) using the following syntax.
```{sql, connection=con}
/* SQL Chunk */
INSERT INTO Customers (PersonID, Name, Street, City, State)
VALUES (5, 'John Smith', '2378 N Main St', 'Flagstaff', 'AZ');
```


### UPDATE a row
Updating a row in the database is similar, but we need to specify which rows to 
update. This is done by including a "WHERE" clause, which feels similar to the 
`dplyr::filter()` function. The only major difference is that SQL uses the single `=`
sign for both assignment and testing equality.

```{sql, connection=con}
/* SQL Chunk */
UPDATE Customers
SET Street = '5638 Daisy Ln', City='Mountainaire', State='AZ'
WHERE PersonID = 3 AND Name='Robert Buscaglia';
```

In the WHERE statement, multiple test conditions must be separated by AND or OR conjunctions.

```{sql, connection=con}
/* SQL Chunk */
SELECT * from Customers
```


## Passing R variables into SQL chunks

Inevitably I would like to be able to create a SQL command that depends on some
dynamic R variable. For example, perhaps I'd like to write a script that queries
the user for a person's name and then prints out all the transactions associated 
with the person. To do this, we need a mechanism to pass an R variable into the
SQL command. 

```{r}
# In an R chunk, I've defined some object that I'd like to use in a SQL statement.
WhichRetailer <- 2
```

```{sql connection=con}
/* Anything with a ? prepended will be                              *
 * replaced with the R object of the same name.                     */
SELECT * FROM Transactions WHERE RetailID = ?WhichRetailer
```


There are some additional situations where a simple character string substitution
doesn't quite work because of how R interprets character strings.

<!-- A quick example of how the `glue` package is intended to work would be helpful here. -->
<!-- The package `glue` is intended make it easy to combine R variables with character -->
<!-- strings. -->

For example, selecting a person by name requires the character string to be quoted, and
some SQL databases have custom quotation syntax. So we'll use the `glue` package to
decide what the appropriate quotation syntax is. The curly brackets tell
`glue` that we want to work with R variable `customers` not the literal string.


```{r}
# R chunk
customer <- 'Derek Sonderegger'                      # For some databases, this works
customer <- glue::glue_sql('{customer}', .con=con)   # This should always work
customer
```

```{sql connection=con}
/* SQL Chunk */
SELECT * FROM Customers WHERE name IN (?customer)
```

If you want to have multiple variables in a SQL statement, we need to tell `glue_sql()`
to make a single SQL command using a `{r_vector*}` notation. The curly brackets tell
`glue` that we want to work with R variable `customers` not the literal string.

```{r}
# R chunk
customers <- c('Derek Sonderegger', 'Aubrey Sonderegger')
glue::glue_sql("{customers}", .con=con)
```
and if I append a `*`, that tells `glue` to make a single SQL statement, not two separate statements.
```{r}
# R Chunk
glue::glue_sql("{customers*}", .con=con)

```

Now we can happily select multiple people from our Customers table.
```{r}
# R Chunk
customer = glue::glue_sql("{customers*}", .con=con)
```

```{sql connection=con}
/* SQL Chunk */ 
SELECT * FROM Customers WHERE name IN (?customer)
```


These SQL statements can be whatever you'd like, and we can happily insert rows
into tables as well.
```{r}
# R Chunk
newCustomer <- c(4, 'Mike Wazowski', '1102 Main St, Apt A113', 'Phoenix', 'AZ')
newCustomer <- glue::glue_sql('{newCustomer*}', .con = con)
```

```{sql connection=con}
/* SQL Chunk */
INSERT INTO Customers (PersonID, Name, Street, City, State)
VALUES(?newCustomer)
```

And this has happily inserted Mike into our `Customers` table.
```{sql, connection=con}
/* SQL Chunk */
SELECT * FROM Customers
```




<!-- ## Best Practices -->

<!-- 1.  Functions should always have a connection object. Production vs Development -->
<!--     as well as different customers database. -->
<!-- 2.  Try to keep as much work done on the database side as possible.  -->
<!--     a)  Typically the computer the database is being run on is WAY faster than your  -->
<!--         client machine. -->
<!--     b) Also, the biggest bottleneck is is usually the network lag associated with  -->
<!--        transferring data from the database, across the internet, and to the client. -->
<!--        Therefore reducing the amount of data passed will be helpful.  -->


## Reading tables using `dbplyr`

There are a lot of good things about SQL, but for database queries, I would 
really like to pretend that the tables are in memory and use all of my favorite 
`dplyr` tools and pipelines. This would mean that I don't have to remember all 
the weird SQL syntax. However, the database interface `dbplyr` is ONLY intended 
for queries and NOT for updating or inserting rows into the tables.

The way this will work is that we will use the previously established database 
connection `con` to create a virtual link between the database table and some 
appropriately named R object.

```{r}
# connect the database tables to similarly named objects in R
Cards <-        tbl(con, 'Cards')
Customers <-    tbl(con, 'Customers')
Retailers <-    tbl(con, 'Retailers')
Transactions <- tbl(con, 'Transactions')
```

However, this does NOT download the whole table into R.  Instead it grabs only a 
couple of rows so that we can see the format. Notice that we don't know how many 
rows are in the Transactions table.
```{r}
Transactions %>% head(3)
# Transactions %>% tail(3)  # not supported because we haven't yet downloaded much information.
```

The guiding principle of `dbplyr` is to delay as much work for as long as possible 
actually pulling the data from the database. The rational is that we spend a great 
deal of time figuring out what the query should look like and too often we write a 
query that accidentally downloads millions of lines of data and slows down our network 
connection. Instead `dbplyr` returns just the first few rows of whatever query we are 
working on until we finish the pipeline with a `collect()` command that will cause 
us to download ALL of the query results and save them as a local `data.frame`.

```{r, warning=FALSE, message=FALSE}
CC_statement <- 
  Customers %>% 
  filter(Name == 'Derek Sonderegger') %>% select(PersonID) %>%
  left_join(Cards) %>% left_join(Transactions) %>% left_join(Retailers) %>%
  select(DateTime, Name, Amount) %>%
  rename(Retailer = Name) 

CC_statement
```

At this point, we *still* haven't downloaded all of the rows. Instead this is 
still a *lazy* query. To actually download everything, we'll pipe this into the 
`collect` function.

```{r}
CC_statement %>%
  collect()
```


<!-- Next have an example where we need to do a mutate command and see when we need -->
<!-- to throw the collect() statement into the pipeline. Emphasize that we want to  -->
<!-- keep the collect() statement as far down the pipeline as possible. -->


It can be fun to see what the SQL code that is being generated is.

```{r}
CC_statement %>% show_query()
```

The algorithm used to convert my `dplyr` statement into a SQL statement doesn't 
mind nesting SQL statements and isn't the same as what I generated by hand, but 
it works.

The last step of a script should be to close the database connection.
```{r}
# Close our database connection when we are through...
dbDisconnect(con)
```

## Exercises {#Exercises_DataBases}
1.  In this exercise, you'll create a database containing the `nycflights13` data. 
    Make sure that you've already downloaded the `nycflights13` package, but DO
    NOT load the `nycflights13` package.
    a. Create a SQLite database and connect to it using the following code: 
        ```{r, eval=FALSE}
        library(dplyr)
        library(dbplyr)
        # Start up a SQL-Lite database with the NYCFlights13 data pre-loaded
        con <- dbplyr::nycflights13_sqlite( )
        ```
    b.  Through the `con` connection object, create links to the `flights` and 
        `airlines` tables using the `tbl` command. DO NOT import the complete 
        `flights` table. We want to delay downloading the data from the database as
        long as possible!
    c.  From the `flights` table, summarize the percent of flights with a 
        departure delayed by more than 10 minutes for each airline. *Hint: make a 
        new column of TRUE/FALSE or 1/0 values that indicate if the flight was late. 
        Then take the average of that column.* Produce a table
        that gives the airline name (not the abbreviation) and the percent of 
        flights that are late.
    d. Using the `dbDisconnect()` command to close the connection `con`. 


2.  I have created a package that contains information for a hypothetical ski pass database
    that could used by AZ Snowbowl. This example is inspired by consulting 
    work that I did with Bridger Bowl just outside of Bozeman, MT. We have 5 tables, `Customers`,
    `PassTypes`, `Passes`, `BlackOutDates` and `PatrolIssues`. After downloading the package
    from GitHub, you should read the documentation for each of these tables. Furthermore,
    there is a function `SkiPasses_sqlite()` that loads the information into a SQLite 
    database.
    ```{r, eval=FALSE}
    devtools::install_github('dereksonderegger/SkiPasses')
    library(SkiPasses)
    con <- SkiPasses_sqlite()  # You might want to use refresh=TRUE here...
    ```
    a)  Go to the GitHub repository for this package at 
        [https://github.com/dereksonderegger/SkiPasses](https://github.com/dereksonderegger/SkiPasses). 
        Where would you expect the code that creates the data tables to live? 
        Where would the documentation for the data tables be? Where is the 
        documentation and code for the function `SkiPasses_sqlite()`? Notice that 
        there is a `SkiPasses.db` file in the `inst/extdata/` directory. Poke 
        around the package and check out the code for the `SkiPasses_sqlite()` 
        function. What does the `refresh=TRUE` option do?
    b)  Run the following code to see where the SQLite database file exists on 
        your computer.
        ```{r, eval=FALSE}
        system.file("extdata", "SkiPasses.db", package = "SkiPasses")
        ```
    c)  Insert a new row into the `Customer` data table for yourself. Also insert 
        a row in the `Passes` table for yourself getting a ski pass for the 
        2020-2021 ski season. *Be sure you are inserting rows into the database
        tables and not a local version you saved in your R environment.*
    d)  Close your database connection and then reopen it. Write a SQL command to 
        verify that you are still in the customer data table.
    e)  Create a function that takes a `PassID` and date and returns either a TRUE or FALSE
        value indicating if the pass authorizes the customer to ski for the day. Your function
        should take the database connection, `PassID`, and `Date` as input parameters. The function
        should ONLY access the database tables through the connection! Try to do as much
        filtering on the database side as possible to reduce the amount of information being
        transferred to your function. Demonstrate
        your function working with both an authorized and unauthorized cases. *Hint: read 
        the documentation of the `Passes` table to understand when a pass is valid. Your code
        will need to identify if there is a valid pass (ie Date is between Start and Finish) and
        assuming there is a valid pass, that `Date` is not one of the blackout dates for that pass.*
    f) Create a graph visualizing the ages of the male and female customers.

    
3.  For this exercise, we'll start a SQLite database and see that the SQLite application 
    stores the data in a very specialized file structure, which usually has a file 
    extension of `.db` or `.sqlite`.
    a. Create the SQLite database file in your current working directory using the following:
        ```{r, eval=FALSE}
        con <- DBI::dbConnect(RSQLite::SQLite(), dbname = "TestSQLiteFile.db")
        
        # Create a table using the iris data
        dbWriteTable(con, 'IRIS', iris)
        
        dbListTables(con)         # What tables are available...
        dbReadTable(con, "IRIS")  # Spit out the IRIS table
        dbDisconnect(con)         # Close connection
        ```
    b)  Now check the files in your current working directory as there should now be a 
        `TestSQLiteFile.db`. The SQLite file structure for data is extremely stable and 
        works across platform types (Unix/Windows, 32/64 bit, big/little endian, etc).  
        As such, it is a good file type choice for storing lots of data in a compact 
        format across different systems (e.g. applications that work on a mobile 
        device vs a computer). While you can open this file using a text editor, you 
        will only see the table declaration of column names and types. The data rows 
        that follow will not be readable.
    c)  Close the file connection and then reconnect to the database. Confirm that
        `IRIS` table is still present in the database.
    d)  When you knit your Rmarkdown document, you might have trouble if you are
        initializing the database each time you knit the document. To get around
        this, you could either run your initialization code once by hand and then
        comment out the initialization steps, or use the `eval=FALSE` code chunk
        option. Comment on solution you chose to do.


